---
title: 'Lecture 5 Module 4: The Special Value `NA`'
subtitle: Exercises
output:
  pdf_document: default
  html_document:
    df_print: paged
---


# Exercises



## Exercise 4.1: Expressions with `NA`

What are the values of these expressions?
```{r}
(7 <= NA) | (12 - 2 > 5) 
```

```{r}
(6 > NA) & (NA == 3) & (7 <= -5)
```

```{r}
(TRUE & NA) | !(FALSE & TRUE)
```

**Solution**



## Exercise 4.2: Working with `NA`

This is a very important technique.

Here's a vector that contains missing data:
```{r}
exercise.4.2.data <-
    c(3, 5, 7, NA, 4, 9, NA, 3)
```

I want to answer three questions about this vector:

* Does the vector contain any missing data?

* How many missing elements does the vector contain?

* What are the locations of the missing data?

Hint: use logical functions and the function `is.na()`.

**Answer**




## Exercise 4.3: Summary statistics

Recall the vector from exercise 4.2:
```{r}
exercise.4.2.data <-
    c(3, 5, 7, NA, 4, 9, NA, 3)
```

Calculate the sum of the non-missing values of this vector.

Then calculate the sample mean of the non-missing values of this vector.

**Solution**



# Solutions to the Exercises


## Exercise 4.1: Expressions with `NA`

What are the values of these expressions?
```{r}
(7 <= NA) | (12 - 2 > 5) 
```

```{r}
(6 > NA) & (NA == 3) & (7 <= -5)
```

```{r}
(TRUE & NA) | !(FALSE & TRUE)
```

**Solution**







## Exercise 4.2: Working with `NA`

This is a very important technique.

Here's a vector that contains missing data:
```{r}
exercise.4.2.data <-
    c(3, 5, 7, NA, 4, 9, NA, 3)
```

I want to answer three questions about this vector:

* Does the vector contain any missing data?

* How many missing elements does the vector contain?

* What are the locations of the missing data?

Hint: use logical functions and the function `is.na()`.

**Answer**

To determine if the vector contains missing data, we have:
```{r}
any( is.na( exercise.4.2.data ) )
```

To count the number of missing elements, we have:
```{r}
sum( is.na( exercise.4.2.data ) )
```

The locations of the missing elements are:
```{r}
which( is.na( exercise.4.2.data ) )
```


## Exercise 4.3: Summary statistics

Recall the vector from problem 4.2:
```{r}
exercise.4.2.data <-
    c(3, 5, 7, NA, 4, 9, NA, 3)
```

Calculate the sum of the non-missing values of this vector.

Then calculate the sample mean of the non-missing values of this vector.

**Solution**

The sum of the non-missing values is:
```{r}
sum( exercise.4.2.data, na.rm = TRUE )
```

The sample mean of the non-missing values is:
```{r}
mean( exercise.4.2.data, na.rm = TRUE )
```





