---
title: 'Lecture 4 Module 2: Iterate vs. Vectorize'
subtitle: Code Examples
output:
  pdf_document: default
  html_document:
    df_print: paged
---


# Section 1: Iteration vs. Vectorization


## Example 1: Counting the elements in animal.vector

For instance, let's go back to the code we used to count the number of elements in `animal.vector`:
```{r}
animal.vector <-
    c( "mouse", "dog", "bison" )

counter <- 0

for( current.animal in animal.vector ) {
    
    counter <- counter + 1
    
}

counter
```


## Example 2: Using the built-in R function length() to count the elements

Of course, we could have just used the built-in R function `length()`:
```{r}
length( animal.vector )
```


## Example 3: Calculating the total sum of the values in a numeric vector

Likewise, let's consider the code for calculating the total sum of the values in a numeric vector:
```{r}
numeric.vector <- 1:5

total <- 0

for( current.number in numeric.vector ) {
    
    total <- total + current.number
}

total
```


## Example 4: Calculating the sum of the values using the built-in R function

We could have just used the `sum()` function:
```{r}
sum( numeric.vector )
```



# Section 2: The Three Principles

## Example 5: Creating two example vectors

Let's create two vectors to represent these variables:
```{r}
x.vector <-
    c( 3, 1, 5 )

y.vector <-
    c( 7, 9, 2 )
```


## Example 6: Calculating the dot product using iteration

Here's the code:
```{r}
running.total <- 0

index.vector <- 1:length( x.vector )

for( index in index.vector ) {
    
    running.total <-
        running.total +
        (x.vector[ index ] * y.vector[ index ])
}

running.total
```


## Example 7: Calculating the dot product using vectorized operations

Now let's calculate the dot product using vectorized operations and the built-in R function `sum()`:
```{r}
sum( x.vector * y.vector )
```


## Example 8: Printing out a simple report

Let's consider our very first `for` loop, where we simply printed out the elements of the `animal.vector`:
```{r}
animal.vector <- c( "mouse", "dog", "bison" )

for( current.animal in animal.vector ) {
    
    cat( "Current animal:", current.animal, "\n" )
    
}
```


## Example 9: Vectorized mapping of a function over a vector

You might think back to this code:
```{r}
exp( 1:5 )
```


## Example 10: Trying to vectorize a cat() statement

Let's try this, but before you run the code try to figure out what it's going to do:
```{r}
cat( "Current animal:", animal.vector, "\n" )
```



# Section 3: Avoid Optimizing



## Example 11: Using a random number generate to populate vectors

I'm going to use a random number generator to populate these vectors:
```{r}
random.vector.1 <-
    rnorm( 100000 )

random.vector.2 <-
    rnorm( 100000 )
```


## Example 12: Calculating the dot product using vectorized operations

First let's calculate the dot product of these two vectors using a vectorized approach:
```{r}
sum( random.vector.1 * random.vector.2 )
```


## Example 13: Calculating the dot product using iteration

Next, let's calculate the dot product using an iterative approach with a `for` loop:
```{r}
total <- 0

index.vector <- 1:length( random.vector.1 )

for( index in index.vector ) {
    
    total <-
        total +
        (random.vector.1[ index ] * 
             random.vector.2[ index ])
}

total
```



# Section 4: Final Grades

Here's an example where iteration is very appealing, even though you don't have to use it.


## Example 14: Creating preliminary score vectors

Let's make some vectors to represent these scores:
```{r}
score.1.vector <-
    c( 85, 92, 95, 82, 91 )

score.2.vector <-
    c( 88, 91, 92, 86, 89 )
```


## Example 15: Using the max() function

Now there is a built-in R function called `max()`, and this takes a set of numbers and returns the maximum value.
```{r}
max( 4, 7 )
```


## Example 16: Using the max() function with two vectors

Here's the code, but before we run it, try to think what's going to happen:
```{r}
max( score.1.vector, score.2.vector )
```


## Example 17: Creating a new storage vector

Here's how we create a numeric vector with storage for 5 elements:
```{r}
graduate.course.score.vector <-
    numeric( 5 )
```


## Example 18: Using iteration to calculate graduate course scores

Now let's construct a `for` loop:
```{r}
for( index in 1:length( score.1.vector ) ) {
    
    graduate.course.score.vector[ index ] <-
        max(
            score.1.vector[ index ],
            score.2.vector[ index ]
        )
}
```



## Example 19: Directly displaying the graduate course scores

Now let's directly display the `graduate.course.score.vector`:
```{r}
graduate.course.score.vector
```
























