---
title: "Lecture 3 Module 3: Constructing Vectors"
subtitle: "Exercises"
output: html_notebook
---



# Exercises


## Exercise 1: Creating a vector

Create a vector with a sequence consisting of three parts:

* First, the value 5 is repeated 10 times.

* Second, the value -2 is repeated 6 times.

* Finally, the value 8 is repeated 3 times.

Report your result using a `cat()` statement, displaying the values with 2 decimal places.

**Solution**



## Exercise 2: The colon operator

Create a vector of the form:
$$
1000,\ 999,\ 998,\ \ldots, 922,\ 921,\ 920
$$

Report the first 10 elements of this vector using a `cat()` statement, displaying the values with 2 decimal places.

**Solution**





## Exercise 3: Creating a vector

Without using the c() function create this vector:
$$
-3.7, -2.7, -1.7, -0.7,\ 0.3,\ 1.3,\ 2.3,\ 3.3,\ 4.3
$$

**Solution**




## Exercise 4: Creating a vector

Use the colon operator twice to create two vectors that are then combined to create the final vector:
$$
\{\ 10,\ 9,\ 8,\ 7,\ 6,\ 5,\ 4,\ 3,\ 2,\ 1,\ 0,\ 1,\ 2,\ 3,\ 4,\ 5,\ 6,\ 7,\ 8,\ 9,\ 10\ \}
$$

**Solution**





## Exercise 5: Constructing a sequence

Construct a sequence that starts at 2.3 and goes to 4.5 in increments of 0.2.

**Solution**





## Exercise 6: Constructing a sequence

Construct a sequence starting at 12.8 and ending at 49.6 of length 200.

**Solution**





## Exercise 7: Odd numbers

Write R code to generate the sequence of the first 15 positive odd numbers.

**Solution**




# Solutions to the Exercises

## Exercise 1: Creating a vector

Create a vector with a sequence consisting of three parts:

* First, the value 5 is repeated 10 times.

* Second, the value -2 is repeated 6 times.

* Finally, the value 8 is repeated 3 times.

Report your result using a `cat()` statement, displaying the values with 2 decimal places.

**Answer**

Here's my solution:
```{r}
exercise.2.1.vector <-
    c(
        rep( x = 5, times = 10 ),
        rep( x = -2, times = 6),
        rep( x = 8, times = 3 )
    )

cat(
    "Exericse 2.1 vector:",
    formatC(
        exercise.2.1.vector,
        format = "f",
        digits = 2
    ),
    fill = TRUE
)
```


## Exercise 2: The colon operator

Create a vector of the form:
$$
1000,\ 999,\ 998,\ \ldots, 922,\ 921,\ 920
$$

Report the first 10 elements of this vector using a `cat()` statement, displaying the values with 2 decimal places.

**Solution**

```{r}
long.vector <-
    1000:920

cat(
    "long.vector:",
    formatC(
        head( long.vector, n = 10 ),
        format = "f",
        digits = 2
    )
)
```



## Exercise 3: Creating a vector

Create a vector of the form:
$$
-3.7, -2.7, -1.7, -0.7,\ 0.3,\ 1.3,\ 2.3,\ 3.3,\ 4.3
$$

**Solution**

Here's one approach:
```{r}
c( -3.7:-0.7, 0.3:4.3 )
```

Here's another:
```{r}
-3.7:4.3
```



## Exercise 4: Creating a vector

Use the colon operator twice to create two vectors that are then combined to create the final vector:
$$
\{\ 10,\ 9,\ 8,\ 7,\ 6,\ 5,\ 4,\ 3,\ 2,\ 1,\ 0,\ 1,\ 2,\ 3,\ 4,\ 5,\ 6,\ 7,\ 8,\ 9,\ 10\ \}
$$

**Solution**

Here's one simple approach:
```{r}
c( 10:0, 1:10 )
```

Here's an alternative way to do this:
```{r}
c(10:1, 0:10 )
```
Really, both methods are fine.





## Exercise 5: Constructing a sequence

Construct a sequence that starts at 2.3 and goes to 4.5 in increments of 0.2.

**Solution**

```{r}
seq( from = 2.3, to = 4.5, by = 0.2 )
```



## Exercise 6: Constructing a sequence

Construct a sequence starting at 12.8 and ending at 49.6 of length 200.

**Solution**

```{r}
seq( from = 12.8, to = 49.6, length.out = 200 )
```


## Exercise 7: Odd numbers

Write R code to generate the sequence of the first 15 odd numbers.

**Solution**

```{r}
seq( from = 1, by = 2, length.out = 15 )
```




