---
title: 'Lecture 2 Module 1: Arithmetic'
subtitle: Exercises
output:
  pdf_document: default
  html_notebook: default
---



# Exercises


## Exercise 1.1: Weird Arithmetic

Consider this method for determining a number:

* We start with the value 17, then we multiply it by 11.

* We then raise the resulting quantity to the third power.

* We then add 1,653 to this value.

Determine if the final value is evenly divisible by 7.

**Solution**

```{r}
# Type your solution here.
```




## Exercise 2.2: Operator Precedence

I recently had to perform a calculation in which I took the three numbers -4, 2, and 10, squared them, multiplied these squared values by the weights 0.25, 0.25, and 0.50, and then added up the resulting values:

| Value |Squared Value | Weight | Squared Value $\times$ Weight |
|----:|------:|--------:|----:|
|  -4 |   16  |   0.25  |  4  |
|   2 |    4  |   0.25  |  1  |
|  10 |  100  |   0.50  | 50  |
|-----|-------|---------|-----|
|     |       | **Sum** |  55 |

I wrote this R code:
```{r}
(-4^2 * 0.25) + (2^2 * 0.25) + (10^2 * 0.50)
```
Why did I get the wrong answer with this expression? 

Then write R code that will perform the calculation correctly.

**Solution**




# Solutions to the Exercises

## Exercise 1.1: Weird Arithmetic

Consider this method for determining a number:

* We start with the value 17, then we multiply it by 11.

* We then raise the resulting quantity to the third power.

* We then add 1,653 to this value.

Determine if the final value is evenly divisible by 7.

**Solution**

I will perform all the arithmetic operations to construct the final number, and then determine whether or not it is divisible by 7 by examinging the remainder when using integer division with a divisor of 7.
```{r}
( (17 * 11)^3 + 1653) %% 7
```

Since the remainder is 0, we conclude that the final number is evenly divisible by 7. To check this, let's first calculate the final number:
```{r}
(17 * 11)^3 + 1653
```

Let's perform integer division on this value:
```{r}
6540856 %/% 7
```

Now let's multiply this value by 7:
```{r}
934408 * 7
```



## Exercise 2.2: Operator Precedence

I recently had to perform a calculation in which I took the three numbers -4, 2, and 10, squared them, multiplied these squared values by the weights 0.25, 0.25, and 0.50, and then added up the resulting values:

| Value |Squared Value | Weight | Squared Value $\times$ Weight |
|----:|------:|--------:|----:|
|  -4 |   16  |   0.25  |  4  |
|   2 |    4  |   0.25  |  1  |
|  10 |  100  |   0.50  | 50  |
|-----|-------|---------|-----|
|     |       | **Sum** |  55 |

I wrote this R code:
```{r}
(-4^2 * 0.25) + (2^2 * 0.25) + (10^2 * 0.50)
```
Why did I get the wrong answer with this expression? 

Then write R code that will perform the calculation correctly.

**Solution**

```{r}
((-4)^2 * 0.25) + (2^2 * 0.25) + (10^2 * 0.50)
```































































