---
title: 'Week 5 Module 6: Histograms and Curves'
subtitle: 'CSCI E-5a: Introduction to R'
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

Let's clear the global environment:
```{r}
rm( list = ls() )
```


And now let's load in the Module 6 R objects:
```{r}
load( "Module 6 R Objects.Rdata" )
```






# Module Overview and Learning Objectives


Hello! And welcome to Module 6: Histograms and Curves.

In this module, we'll examine methods for superimposing curves on histograms.



* In section 1, we'll review the theory of the normal distribution.

* In section 2, we'll see how to superimpose a density curve over a histogram.

* In section 3, we'll learn how to determine the best-fitting density curve for a normally distributed set of data.

* In section 4, we'll explore non-parametric density curves.


When you've completed this module, you should be able to:

* Superimpose a normal density curve over a histogram.

* Determine the parameters for the best-fitting normal density curve.

* Superimpose a non-parametric density curve over a histogram.

* Draw non-parametric empirical density curves with normal density curves or other non-parametric density curves.


There are two new built-in R functions in this module:

* `dnorm()`

* `density()`



All right! Let's get started with a review of the normal distribution.





# Section 1: Review of the Normal Distribution

> **Main Idea:** *Let's review the normal distribution*

In this section, we'll explore the theory of the normal distribution.



Much of classical statistics is based on objects called *normal distributions*.

A *parameter* of a distribution is a value that must be specified in order to completely determine a normal distribution.

A normal distribution has two parameters:

* The first parameter, denoted by the Greek letter $\mu$, is the *mean* of the distribution, which describes the central location of the density curve.

* The second parameter, denoted by the Greek letter $\sigma$, is the *standard deviation* of the distribution, which describes how spread out the distribution is.

Once we've specified the mean and standard deviation for a normal distribution, we've completely determined it.



Every normal distribution has an associated "density curve".

Normal density curves have a characteristic symmetric "mountain" shape, where the center of the mountain is determined by the value of the mean $\mu$, and the width of the mountain is determined by the value of the standard deviation $\sigma$.



The built-in R function `dnorm()` can calculate normal density curves.

The `dnorm()` function takes three input arguments:

* The first input argument is `x`, and this is just the value at which the density function is to be evaluated.

* The second input argument is `mean`, and this is the mean of the normal distribution.

* The third input argument is `sd`, and this is the standard deviation of the normal distribution.

The `dnorm()` function returns the value of the normal density function at the specified value of `x`, for the normal distribution with population mean `mean` and population standard deviation `sd`.


Let's make a graph of the density function for the normal distribution, with mean $\mu = 100$ and standard deviation $\sigma = 10$:
```{r}
curve(
    expr = dnorm(x, mean = 100, sd = 10),
    main = "Graph of normal density curve",
    xlim = c(70, 130),
    ylim = c(0, 0.05),
    xlab = "x",
    ylab = "Density",
    lty = "solid",
    lwd = 3,
    col = "black"
)


segments(
    x0 = 70,
    y0 = 0,
    x1 = 130,
    y1 = 0,
    lty = "solid",
    lwd = 2,
    col = "gray60"
)

```


So that's a quick overview of the theory of the normal distribution.

Now let's see how to superimpose a mathematical curve over a histogram.




## Exercise 6.1: Drawing a normal density curve

Construct a graph of the density function for the normal distribution, with mean $\mu = 45$ and standard deviation $\sigma = 7$.

**Solution**





# Section 2: Superimposing a Normal Density Curve

> **Main Idea:** *We can superimpose a density curve on a histogram*

In this section, we'll see how to superimpose a density curve over a histogram.


Let's make a histogram of the values in the `example.6.1.data` vector:
```{r}
hist(
    x = example.6.1.data,
    main = "Histogram of random normal values",
    col = "cornsilk",
    xlab = "x",
    ylab = "y",
    breaks = 50
)
```

This data looks like it might come from a normal distribution.

To help us decide if this data comes from a normal distribution, we will *superimpose* the density curve of a particular normal distribution on top of the histogram of the observed data:

* First, we create the histogram using the `hist()` function.

* Next, we draw the density curve on top of the histogram using the `curve()` function.



In order to do this, we need to set two options:

* In the `hist()` function, we need to set `prob = TRUE`.

* In the `curve()` function, we need to set `add = TRUE`.

The `prob = TRUE` option in the `hist()` function means that the histogram will display the density on the $y$ axis.

If you don't do this, the density curve will appear as a flat line along the $x$-axis.

The `add = TRUE` option in the `curve()` function means that the curve will be drawn on the existing plot instead of creating a new plot.

Let's redraw the histogram of `example.6.1.data`, and this time we'll superimpose a normal density curve with a mean of 100 and a standard deviation of 10:
```{r}
hist(
    x = example.6.1.data,
    prob = TRUE,
    main = "Histogram of random normal data",
    xlab = "x",
    ylab = "Density",
    col = "cornsilk",
    breaks = 50
)

curve(
    dnorm(
        x,
        mean = 100,
        sd = 10
    ),
    lty = "dashed",
    lwd = 4,
    col = "darkred",
    add = TRUE
)
```



So that's how to superimpose a normal density curve over a histogram.

Now let's see how to fit a normal density curve to data.





## Exercise 6.2: Superimposing a normal density curve

Draw the histogram of the data in `exercise.6.2.data`. Then superimpose a normal density curve with a mean of 45 and a standard deviation of 7:

**Solution**





# Section 3: Fitting a Normal Density Curve

> **Main Idea:** *We can fit a normal density curve to data*

In this section, we'll learn how to determine the best-fitting density curve for a normally distributed set of data.



How do we choose the values of the `mean` and `sd` parameters when we fit a normal distribution to this data?

According to statistical theory, the best normal distribution to choose to fit this data is the normal distribution that has a mean equal to the observed sample mean, and a standard deviation equal to the observed sample standard deviation.


```{r}
hist( 
    example.6.1.data,
    main = "Histogram of values in normal.data.set",
    col = "cornsilk",
    xlab = "x",
    ylab = "y",
    prob = TRUE,
    breaks = 50
)


sample.mean <-
    mean( example.6.1.data )

sample.standard.deviation <-
    sd( example.6.1.data )

curve(
    dnorm(
        x,
        mean = sample.mean,
        sd = sample.standard.deviation
    ),
    lty = "dashed",
    lwd = 3,
    col = "darkred",
    add = TRUE
)
```


Notice that this method *requires* that the population is normally distributed; the only question is to determine the best-fitting curve from among all possible normal distributions.

If the population is not normally distributed, then this method won't make much sense.

To see an example of this, let's make a histogram of `example.6.2.data`:
```{r}
hist(
    example.6.2.data,
    main = "Histogram of non-normal random values",
    col = "pink",
    xlab = "x",
    ylab = "y",
    prob = TRUE,
    breaks = 50
)
```

This doesn't have the familiar mountain shape of the normal distribution.

Now we'll attempt to fit a normal curve to this data by using the observed sample mean and sample variance values:
```{r}
hist(
    example.6.2.data,
    xlim = c(60, 80),
    main = "Histogram of random non-normal values",
    col = "pink",
    xlab = "x",
    ylab = "y",
    prob = TRUE,
    breaks = 50
)

sample.mean <-
    mean( example.6.2.data )

sample.standard.deviation <-
    sd( example.6.2.data )

curve(
    dnorm(
        x,
        mean = sample.mean,
        sd = sample.standard.deviation
    ),
    lty = 2,
    lwd = 3,
    col = "black",
    add = TRUE
)
```

This didn't work very well at all.

The conclusion here is that you shouldn't try to fit a normal density curve to data that isn't normally distributed.



So that's how to fit a normal density curve to data.

Now let's see how to relax the assumption of normality, and construct density curves for non-normal data.




## Exercise 6.3: Fitting a normal density curve

Calculate the sample mean and the sample standard deviation of the values in `exercise.6.2.data`. Redraw the histogram, and fit a normal density curve using the sample mean and sample standard deviation.

**Solution**





# Section 4: Fitting a Kernel Density Curve

> **Main Idea:** *We can fit a non-parametric kernel density curve to data*

In this section, we'll explore non-parametric kernel density curves.


When we fit a normal density curve to data, we are assuming that the data follows a normal distribution, and we just have to determine the mean and standard deviation to use.

A *kernel density estimator* doesn't make any assumption of normality, or any other distribution -- it just looks at the data and tries to create a curve that becomes large when there are many data points in a region, and becomes small when there are few data points in a region.

Sometimes I'll use the terms such as "empirical density curve" or "non-parametric density function", but this means the same thing as "kernel density curve".

We can fit a kernel density curve to the data in `example.6.2.data` using the `density()` function.

We can then plot this using the `lines()` function (no need to use `add = TRUE`).
```{r}
hist(
    example.6.2.data,
    main = "Histogram of values in non.normal.data.set",
    col = "pink",
    xlab = "x",
    ylab = "y",
    prob = TRUE,
    breaks = 50
)

lines(
    density( example.6.2.data ),
    lty = "dotted",
    lwd = 4,
    col = "darkorchid3"
)
```
That works well!

The kernel density estimator works really well with normal data:
```{r}
hist(
    example.6.1.data,
    main = "Histogram of random normal values",
    col = "lightcyan2",
    xlab = "x",
    ylab = "y",
    prob = TRUE,
    breaks = 50
)

lines(
    density( example.6.1.data ),
    lty = "dashed",
    lwd = 3,
    col = "black"
)
```

We can dispense with the histogram entirely, and just draw the best-fitting normal density along with the non-parametric empirical density:
```{r}
plot(
    x = NULL,
    xlim = c(70, 130),
    ylim = c(0, 0.05),
    main =
        "Comparing best-fit normal and empirical densities",
    xlab = "x",
    ylab = "Density"
)


segments(
    x0 = 70,
    y0 = 0,
    x1 = 130,
    y1 = 0,
    lty = "solid",
    lwd = 2,
    col = "gray50"
)

sample.mean <-
    mean( example.6.1.data )

sample.standard.deviation <-
    sd( example.6.1.data )

curve(
    dnorm(
        x, 
        mean = sample.mean, 
        sd = sample.standard.deviation
    ),
    lty = 1,
    lwd = 3,
    col = "dodgerblue3",
    add = TRUE
)

lines(
    density( example.6.1.data ),
    lty = 2,
    lwd = 4,
    col = "navy"
)
```

Empirical density curves can be very useful when you want to visualize the distribution of multiple variables in one display.
```{r}
plot(
    x = NULL,
    xlim = c(80, 160),
    ylim = c(0, 0.05),
    main = "Comparing two empirical density functions",
    xlab = "x",
    ylab = "Density"
)


segments(
    x0 = 80,
    y0 = 0,
    x1 = 160,
    y1 = 0,
    lty = "solid",
    lwd = 2,
    col = "gray50"
)



lines(
    density( example.6.3.data ),
    lty = "solid",
    lwd = 3,
    col = "blue"
)

lines(
    density( example.6.4.data ),
    lty = "solid",
    lwd = 3,
    col = "red"
)


legend(
    x = 80,
    y = 0.05,
    legend =
        c( "Group A", "Group B" ),
    lty = "solid",
    lwd = 3,
    col =
        c( "blue", "red" )
)

```


So that's how to fit a non-parametric kernel density curve to data.

Now let's review what we've learned in this module.




## Exercise 5.4: Superimposing a non-parametric density curve

Construct a histogram of the values in `exercise.6.4.data`, and then superimpose a normal density curve. Finally, superimpose a non-parameteric density curve over this display.

**Solution**






# Module Review

In this module, we examined methods for superimposing curves on histograms.


* In section 1, we reviewed the theory of the normal distribution.

* In section 2, we saw how to superimpose a density curve over a histogram.

* In section 3, we learned how to determine the best-fitting density curve for a normally distributed set of data.

* In section 4, we explored non-parametric density curves.



Now that you've completed this module, you should be able to:

* Superimpose a normal density curve over a histogram.

* Determine the parameters for the best-fitting normal density curve.

* Superimpose a non-parametric density curve over a histogram.

* Draw non-parametric empirical density curves with normal density curves or other non-parametric density curves.



There were two new built-in R functions in this module:

* `dnorm()`

* `density()`



All right! That's it for Module 6: Histograms and Curves.

In fact, that's all the content for Week 5: Logical Values.

Now you can finish Problem Set 5.





# Solutions to the Exercises


## Exercise 6.1: Drawing a normal density curve

Construct a graph of the density function for the normal distribution, with mean $\mu = 45$ and standard deviation $\sigma = 7$.

**Solution**

```{r}
curve(
    dnorm(x, mean = 45, sd = 7),
    main = "Graph of normal density curve",
    xlim = c(20, 70),
    ylim = c(0, 0.06),
    xlab = "x",
    ylab = "Density",
    lty = "solid",
    lwd = 3,
    col = "black"
)


segments(
    x0 = 70,
    y0 = 0,
    x1 = 130,
    y1 = 0,
    lty = "solid",
    lwd = 2,
    col = "gray60"
)

```



## Exercise 6.2: Superimposing a normal density curve

Draw the histogram of the data in `exercise.6.2.data`. Then superimpose a normal density curve with a mean of 45 and a standard deviation of 7:

**Solution**

```{r}
hist(
    exercise.6.2.data,
    prob = TRUE,
    main = "Histogram of random normal data",
    xlab = "x",
    ylab = "Density",
    col = "cornsilk",
    breaks = 50
)

curve(
    dnorm(
        x,
        mean = 45,
        sd = 7
    ),
    lty = "dashed",
    lwd = 4,
    col = "darkred",
    add = TRUE
)
```



## Exercise 6.3: Fitting a normal density curve

Calculate the sample mean and the sample standard deviation of the values in `exercise.6.2.data`. Redraw the histogram, and fit a normal density curve using the sample mean and sample standard deviation.

**Solution**

```{r}
hist( 
    exercise.6.2.data,
    main = "Histogram of values in exercise.6.2.data",
    col = "skyblue",
    xlab = "x",
    ylab = "y",
    prob = TRUE,
    breaks = 50
)


sample.mean <-
    mean( exercise.6.2.data )

sample.standard.deviation <-
    sd( exercise.6.2.data )

curve(
    dnorm(
        x,
        mean = sample.mean,
        sd = sample.standard.deviation
    ),
    lty = "dashed",
    lwd = 3,
    col = "black",
    add = TRUE
)
```




## Exercise 6.4: Superimposing a non-parametric density curve

Construct a histogram of the values in `exercise.6.4.data`, and then superimpose a normal density curve. Finally, superimpose a non-parameteric density curve over this display.

**Solution**

```{r}
hist( 
    exercise.6.4.data,
    main = "Histogram of values in normal.data.set",
    col = "azure1",
    xlab = "x",
    ylab = "y",
    prob = TRUE,
    breaks = 50
)


sample.mean <-
    mean( exercise.6.4.data )

sample.standard.deviation <-
    sd( exercise.6.4.data )

curve(
    dnorm(
        x,
        mean = sample.mean,
        sd = sample.standard.deviation
    ),
    lty = "dashed",
    lwd = 3,
    col = "royalblue4",
    add = TRUE
)

lines(
    density( exercise.6.4.data ),
    lty = "solid",
    lwd = 2,
    col = "black"
)
```











