---
title: 'Week 6 Module 4: Barplots'
subtitle: Exercises
output:
  pdf_document: default
  html_document:
    df_print: paged
---


Let's clear the global environment:
```{r}
rm( list = ls() )
```

Now let's load in the Module 4 R objects:
```{r}
load( "Module 4 R Objects.Rdata" )
```



# Exercises

## Exercise 1: Simple barplot

Here are the number of support calls for each of the regional offices of WiDgT:

| Location       | Number of support calls |
|:---------------|:-----------------------:|
| Boston         |             43          |
| London         |             27          |
| Salt Lake City |             11          |
| Shanghai       |             56          |

Construct a barplot that displays the frequency counts of the number of support calls across the locations.

**Solution**





## Exercise 2: Barplots and tables

Let's return to the `one.week.cereal.brand.character.string.vector` from the last module:
```{r}
head( one.week.cereal.brand.character.string.vector )
```

First, construct a table of the frequency counts.

Then create a properly formatted barplot for this table.

Finally, construct a table of relative frequencies for this data, and make another properly formatted barplot.

**Solution**





## Exercise 3: Barplots and factors

In the previous exercise, we created a barplot of the frequency counts for the one-week cereal sales.

However, the order of the categories wasn't our usual one, and it would be nice to display the data with our standard conventions.

First, construct a factor where the levels are ordered with Sugar Bomz first, then Krispee Yummm!!, and finally Healthy Kale and Tofu. Also, change the labels of the levels so that they are "SBZ", "KYM", and "HKT", respectively.

Next, construct a frequency count table using this factor, and save this in a variable.

Then create a properly formatted barplot for this table.

**Solution**






# Solutions to the Exercises

## Exercise 1: Simple barplot

Here are the number of support calls for each of the regional offices of WiDgT:

| Location       | Number of support calls |
|:---------------|:-----------------------:|
| Boston         |             43          |
| London         |             27          |
| Salt Lake City |             11          |
| Shanghai       |             56          |

Construct a barplot that displays the frequency counts of the number of support calls across the locations.

**Solution**

```{r}
support.calls.vector <-
  c( 43, 27, 11, 56 )

barplot(
  height = support.calls.vector,
  ylim = c(0, 60),
  main = "Support calls across regional offices",
  xlab = "Locations",
  ylab = "Frequency count",
  col = "lightsteelblue2",
  names.arg = c( "Bos", "Lon", "SLC", "Sha" ),
)
```




## Exercise 2: Barplots and tables

Let's return to the `one.week.cereal.brand.character.string.vector` from the last module:
```{r}
head( one.week.cereal.brand.character.string.vector )
```

First, construct a table of the frequency counts for this data.

Then create a properly formatted barplot for this table.

**Solution**

```{r}
cereal.brand.frequency.count.table <-
  table( one.week.cereal.brand.character.string.vector )

barplot(
  height = cereal.brand.frequency.count.table,
  main = "One-Week Cereal Brand Sales",
  xlab = "Cereal brand",
  ylab = "Frequency count",
  col = "azure2"
)
```



## Exercise 3: Barplots and factors

In the previous exercise, we created a barplot of the frequency counts for the one-week cereal sales.

However, the order of the categories wasn't our usual one, and it would be nice to display the data with our standard conventions.

First, construct a factor where the levels are ordered with Sugar Bomz first, then Krispee Yummm!!, and finally Healthy Kale and Tofu. Also, change the labels of the levels so that they are "SBZ", "KYM", and "HKT", respectively.

Next, construct a frequency count table using this factor, and save this in a variable.

Then create a properly formatted barplot for this table.

**Solution**

First, we'll construct the factor:
```{r}
one.week.cereal.brand.factor <-
  factor(
    x = one.week.cereal.brand.character.string.vector,
    levels = c( "Sugar Bomz", "Krispee Yummm!!", "Healthy Kale and Tofu" ),
    labels = c( "SBZ", "KYM", "HKT" )
  )
```

Next, we'll create the frequency count table, and save this in a variable:
```{r}
one.week.cereal.brand.frequency.count.table <-
  table( one.week.cereal.brand.factor )

one.week.cereal.brand.frequency.count.table
```

Now we can create the barplot:
```{r}
barplot(
  height = one.week.cereal.brand.frequency.count.table,
  ylim = c(0, 200),
  main = "Barplot of one-week sales across cereal brands",
  xlab = "Cereal brands",
  ylab = "Number of boxes sold",
  col = "darkseagreen2"
)
```




