---
title: 'Lecture 5 Module 3: Logical Indexing'
subtitle: Exercises
output:
  pdf_document: default
  html_document:
    df_print: paged
---



# Exercises


## Exercise 3.1: Logical indexing

Construct a logical indexing vector to select the elements "Cat", "Mouse", and "Hedgehog" from `character.vector`.

**Solution**




## Exercise 3.2: Using a comparison operation

The `rivers` vector is a built-in vector consisting of the lengths of the longest 141 rivers in North America.

Select the lengths of the rivers that are strictly longer than 1000 miles.

**Solution**



## Exercise 3.3: Brand-specific sum and mean

We have a small dataset of 7 transactions, and for each transaction we have the brand of cereal and the number of boxes that were purchased:

| Transaction | Brand Name | Number of Boxes |
|:-----------:|:----------:|:---------------:|
|      1      |     SBZ    |        4        |
|      2      |     SBZ    |        6        |
|      3      |     KYM    |        3        |
|      4      |     SBZ    |        5        |
|      5      |     HKT    |        1        |
|      6      |     KYM    |        2        |
|      7      |     HKT    |        2        |

Let's put the brand and number of boxes sold into vectors:
```{r}
brand.name.vector <-
    c( "SBZ", "SBZ", "KYM", "SBZ", 
       "HKT", "KYM", "HKT" )

number.of.boxes.sold.vector <-
    c( 4, 6, 3, 5, 1, 2, 2 )
```


Using logical indexing, determine the total number of boxes of Sugar Bomz sold, and also the average number of boxes sold per transaction.

**Solution**





# Solutions to the Exercises

## Exercise 3.1: Logical indexing

Let's construct a vector of character strings:
```{r}
character.vector <-
    c( "Red", "Cat", "Dog", "Pen", "Phone",
       "White", "Mouse", "Bird", "Piano", "Bob",
       "Anita", "Hedgehog" )
```


Construct a logical indexing vector to select the elements "Cat", "Mouse", and "Hedgehog" from `character.vector`.

**Solution**

For your reference, here's the `character.vector`:
```{r}
character.vector
```

Now we'll construct this logical vector:
```{r}
logical.indexing.vector <-
    c( FALSE, TRUE, FALSE, FALSE,
       FALSE, FALSE, TRUE, FALSE,
       FALSE, FALSE, FALSE, TRUE )
```

Now we'll perform logical indexing:
```{r}
character.vector[ logical.indexing.vector ]
```



## Exercise 3.2: Using a comparison operation

The `rivers` vector is a built-in vector consisting of the lengths of the longest 141 rivers in North America.

Using logical indexing with a vectorized comparison to construct a vector with the values of the lengths of the rivers that are strictly longer than 1000 miles.

**Solution**

```{r}
rivers[ rivers > 1000 ]
```



## Exercise 3.3: Brand-specific sum and mean

We have a small dataset of 7 transactions, and for each transaction we have the brand of cereal and the number of boxes that were purchased:

| Transaction | Brand Name | Number of Boxes |
|:-----------:|:----------:|:---------------:|
|      1      |     SBZ    |        4        |
|      2      |     SBZ    |        6        |
|      3      |     KYM    |        3        |
|      4      |     SBZ    |        5        |
|      5      |     HKT    |        1        |
|      6      |     KYM    |        2        |
|      7      |     HKT    |        2        |

Let's put the brand and number of boxes sold into vectors:
```{r}
brand.name.vector <-
    c( "SBZ", "SBZ", "KYM", "SBZ", 
       "HKT", "KYM", "HKT" )

number.of.boxes.sold.vector <-
    c( 4, 6, 3, 5, 1, 2, 2 )
```


Using logical indexing, determine the total number of boxes of Sugar Bomz sold, and also the average number of boxes sold per transaction.

**Solution**

```{r}
sbz.logical.indexing.vector <-
    brand.name.vector == "SBZ"

sbz.number.of.boxes.sold.vector <-
    number.of.boxes.sold.vector[
        sbz.logical.indexing.vector
    ]

sbz.total.boxes.sold <-
    sum( sbz.number.of.boxes.sold.vector )

cat(
    "Total number of boxes of Sugar Bomz sold:",
    formatC(
        sbz.total.boxes.sold,
        format = "f",
        digits = 2
    )
)
```

Now for the mean number of boxes of Sugar Bomz sold per transaction:
```{r}
sbz.mean.boxes.sold <-
    mean( sbz.number.of.boxes.sold.vector )

cat(
    "Mean number of boxes of Sugar Bomz sold:",
    formatC(
        sbz.mean.boxes.sold,
        format = "f",
        digits = 2
    )
)
```






































































