---
title: 'Lecture 5 Module 2: Sample Counts and Proportions'
subtitle: Exercises
output:
  html_document:
    df_print: paged
  pdf_document: default
---


# Exercises


## Exercise 2.1: Sample sum and proportion

Here's the vector `second.logical.vector`:
```{r}
second.logical.vector <- 
    c(TRUE, FALSE, TRUE, TRUE, FALSE)
```

Using the `sum()` function, determine the total number of elements of `second.logical.vector` that have the value `TRUE`.

Using the `mean()` function, determine the proportion of elements of `second.logical.vector` that have the value `TRUE`.

**Solution**




## Exercise 2.2: Sample sum and proportion

Consider this numeric vector:
```{r}
second.numeric.vector <-
    c( -8.4, 7.3, 2.5, 0.1, -4.4, 6.3, 1.9 )
```

How many elements of this vector are greater than or equal to 4.2?

What proportion of elements of this vector are strictly less than 1.8?

**Solution**






# Solutions to the Exercises


## Exercise 2.1: Sample sum and proportion

Here's the vector `second.logical.vector`:
```{r}
second.logical.vector <- 
    c(TRUE, FALSE, TRUE, TRUE, FALSE)
```

Using the `sum()` function, determine the number of elements of `second.logical.vector` that have the value `TRUE`.

Using the `mean()` function, determine the proportion of elements of `second.logical.vector` that have the value `TRUE`.

**Solution**

To count the number of elements of the vector that have the value `TRUE`, we use the `sum()` function:
```{r}
sum( second.logical.vector )
```

To determine the proportion of elements of the vector that have the value `TRUE`, we use the `mean()` function:
```{r}
mean( second.logical.vector )
```



## Exercise 2.2: Sample sum and proportion

Consider this numeric vector:
```{r}
second.numeric.vector <-
    c( -8.4, 7.3, 2.5, 0.1, -4.4, 6.3, 1.9 )
```

How many elements of this vector are greater than or equal to 4.2?

What proportion of elements of this vector are strictly less than 1.8?

**Solution**

How many elements of this vector are greater than or equal to 4.2?
```{r}
sum( second.numeric.vector >= 4.2 )
```


What proportion of elements of this vector are strictly less than 1.8?
```{r}
mean( second.numeric.vector < 1.8 )
```




