---
title: 'Week 1 Modulue 2: Graphics Plots'
subtitle: Exercises
output:
  pdf_document: default
  html_notebook: default
---


First, let's clear the global computing environment:
```{r}
rm( list = ls() )
```



# Exercises for Week 1 Module 2: Graphics Plots



## Exercise 2.1: Modifying a basic plot

Plot a single point at the location (60, 240). Set the $x$-axis to range between 0 and 100, and the $y$-axis to range between 0 and 400. Be sure to explicitly specify a main title, as well as titles for the $x$- and $y$ axes.

**Solution**

```{r}
# Type your answer here
```



## Exercise 2.2: Using graphical parameters

Plot a single point at the location (60, 90).

Set both the $x$-axis and the $y$-axis to range between 0 and 100, and use a square graph shape.

Set a graphical parameter to remove the box around the display.

Be sure to include all the fancy stuff!

**Solution**

```{r}
# Type your answer here
```



## Exercise 2.3: Creating an empty plot with no data

Create an empty graph with no data where $x$ ranges from 0 to 20 and $y$ ranges from 0 to 5.

**Solution**

```{r}
# Type your answer here
```




## Exercise 2.4: Creating a completely blank display

Create a completely blank display where $x$ ranges from 0 to 20 and $y$ ranges from 0 to 5.

**Solution**




# Solutions to the Exercises


## Exercise 2.1: Modifying a basic plot

Plot a single point at the location (60, 240).

Set the $x$-axis to range between 0 and 100, and the $y$-axis to range between 0 and 400.

Be sure to put in all the extra stuff!

**Solution**

Here's my solution:
```{r}
plot(
    x = 60,
    y = 240,
    xlim = c(0, 100),
    ylim = c(0, 400),
    main = "Plot of a single point",
    xlab = "x",
    ylab = "y",
    las = 1
)
```




## Exercise 2.2: Using graphical parameters

Plot a single point at the location (60, 90).

Set both the $x$-axis and the $y$-axis to range between 0 and 100, and use a square graph shape.

Set a graphical parameter to remove the box around the display.

Be sure to include all the fancy stuff!

**Solution**

Here's my solution:
```{r}
par( bty = "n" )

par( pty = "s" )

plot(
    x = 60,
    y = 90,
    xlim = c(0, 100),
    ylim = c(0, 100),
    main = "Plot of a single point",
    xlab = "x",
    ylab = "y",
    las = 1
)
```


## Exercise 2.3: Creating an empty plot with no data

Create an empty graph with no data where $x$ ranges from 0 to 20 and $y$ ranges from 0 to 5.

**Solution**

Here's my solution:
```{r}
plot(
    x = NULL,
    xlim = c(0, 20),
    ylim = c(0, 5),
    main = "Empty plot",
    xlab = "x",
    ylab = "y",
    las = 1
)
```



## Exercise 2.4: Creating a completely blank display

Create a completely blank plotting region where $x$ ranges from 0 to 20 and $y$ ranges from 0 to 5.

**Solution**

```{r}
plot(
    x = NULL, 
    xlim = c(0, 20), 
    ylim = c(0, 5),
    xlab = "",
    ylab = "",
    axes = FALSE
)
```













