---
title: "Lecture 2 Module 3 Exercises"
output:
  pdf_document: default
  html_notebook: default
---



# Exercise 2.1: Ted Williams Batting Statistics

Here are Ted Williams's career statistics:

| Statistics        | Value |
|:------------------|------:|
| Plate appearances | 9,792 |
| At-bats           | 7,706 |
| Hits              | 2,654 |
| Doubles           |   525 |
| Triples           |    71 |
| Home Runs         |   521 |
| Bases on balls    | 2,021 |
| Hit by a pitch    |    39 |
| Sacrifice flies   |    20 |



## Part (a)

Calculate Ted Williams' career batting average.


## Part (b)

Calculate Ted Williams' career on-base percentage.


## Part (c)

Calculate Ted Williams' career slugging average.



# Solution


Here are Ted Williams's career statistics:

| Statistics      | Value |
|:----------------|------:|
| At-bats         | 7,706 |
| Hits            | 2,654 |
| Doubles         |   525 |
| Triples         |    71 |
| Home Runs       |   521 |
| Bases on balls  | 2,021 |
| Hit by a pitch  |    39 |
| Sacrifice flies |    20 |


## Part (a)

Calculate Ted Williams' career batting average


**Solution**

Let's start by creating a set of variables to hold Williams' career batting statistics:
```{r}
ted.williams.at.bats <- 7706

ted.williams.hits <- 2654

ted.williams.doubles <- 525

ted.williams.triples <- 71

ted.williams.home.runs <- 521

ted.williams.bases.on.balls <- 2021

ted.williams.hit.by.a.pitch <- 39

ted.williams.sacrifice.flies <- 20
```



Now let's calculate the batting average for Ted Williams:
```{r}
ted.williams.batting.average <- ted.williams.hits / ted.williams.at.bats

cat( "Ted Williams batting average:",
     formatC(
         ted.williams.batting.average,
         format = "f",
         digits = 3
     )
)
```


## Part (b)

Calculate Ted Williams' career on-base percentage

**Solution**

Next, let's calculate Ted Williams's on-base percentage:
```{r}
ted.williams.on.base.percentage <-
    (ted.williams.hits +
         ted.williams.bases.on.balls +
         ted.williams.hit.by.a.pitch) /
    (ted.williams.at.bats +
         ted.williams.bases.on.balls +
         ted.williams.hit.by.a.pitch +
         ted.williams.sacrifice.flies)

cat( "Ted Williams on-base percentage:",
     formatC(
         ted.williams.on.base.percentage,
         format = "f",
         digits = 3
     )
)
```


## Part (c)

Calculate Ted Williams' career slugging percentage.

**Solution**

Remember that we have to first compute the number of singles before we can calculate the total bases:
```{r}
ted.williams.singles <-
    ted.williams.hits -
    ted.williams.doubles -
    ted.williams.triples -
    ted.williams.home.runs

ted.williams.total.bases <-
    ted.williams.singles +
    2 * ted.williams.doubles +
    3 * ted.williams.triples +
    4 * ted.williams.home.runs
```


Finally, we can calculate Ted Williams's slugging percentage:
```{r}
ted.williams.slugging.percentage <-
    ted.williams.total.bases / ted.williams.at.bats

cat( "Ted Williams slugging percentage:",
     formatC(
         ted.williams.slugging.percentage,
         format = "f",
         digits = 3
     )
)
```












