---
title: 'Lecture 3, Module 3: Vector Indexing'
subtitle: Exercises
output:
  html_document:
    df_print: paged
---




# Exercises


## Exercise 3.1: Indexing with positive integers

Let's construct a simple vector:
```{r}
exercise.vector <-
    c( 6, 3, 2, 8, 6, 4, 1, 5, 9, 0)
```

### Part (a)

Select the fourth element of this vector.

**Solution**


### Part (b)

Select the elements in the second, fifth, and seventh positions.

**Solution**


### Part (c)

Select the fourth through the eighth elements of the vector.

**Solution**




## Exercise 3.2: Indexing with negative integers

Let's construct a simple vector:
```{r}
exercise.vector <-
    c( 6, 3, 2, 8, 6, 4, 1, 5, 9, 0)
```

### Part (a)

Select all the elements of this vector except for the seventh element.

**Solution**




### Part (b)

Select all the elements of this vector except for the elements in the third, sixth, and eighth positions.

**Solution**


### Part (c)

Select all the elements of the vector except for the first through third elements.

**Solution**



# Solutions to the Exercises

## Exercise 3.1: Indexing with positive integers

Let's construct a simple vector:
```{r}
exercise.vector <-
    c( 6, 3, 2, 8, 6, 4, 1, 5, 9, 0)
```

### Part (a)

Select the fourth element of this vector.

**Solution**

```{r}
exercise.vector[ 4 ]
```



### Part (b)

Select the elements in the second, fifth, and seventh positions.

**Solution**

```{r}
exercise.vector[ c(2, 5, 7) ]
```



### Part (c)

Select the fourth through the eighth elements of the vector.

**Solution**

```{r}
exercise.vector[ 4:8 ]
```



## Exercise 3.2: Indexing with negative integers

Let's construct a simple vector:
```{r}
exercise.vector <-
    c( 6, 3, 2, 8, 6, 4, 1, 5, 9, 0)
```

### Part (a)

Select all the elements of this vector except for the seventh element.

**Solution**

```{r}
exercise.vector[ -7 ]
```


### Part (b)

Select all the elements of this vector except for the in the third, sixth, and eighth positions.

**Solution**

```{r}
exercise.vector[ c( -3, -6, -8 ) ]
```


### Part (c)

Select all the elements of the vector except for the first through third elements.

**Solution**

```{r}
exercise.vector[ -1:-3 ]
```





















































