---
title: 'Week 5 Module 1: Logical Values'
subtitle: 'CSCI E-5a: Programming in R'
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---


Let's clear the global environment:
```{r}
rm( list = ls() )
```


# Module Overview and Learning Outcomes


Hello! And welcome to Module 1: Logical Values.

In this module, we'll study logical values.



* In section 1, we'll define the concept of a logical value.

* In section 2, we'll define three important logical operators, and see how to use them in a vectorized approach.

* In section 3, we'll learn how to generate logical vectors from other types of vectors, in particular numeric vectors.

* In section 4, we'll meet three important logical functions.

* In section 5, we'll discuss some technical issues that arise when comparing floating point numbers.



At the end of this module, you should be able to:

* Explain the concept of logical data.

* Perform logical operations on logical values and vectors.

* Construct logical vectors from other types of vectors, especially numeric vectors.

* Explain how the logical functions `all()`, `any()`, and `which()` operate.

* Explain how to compare floating point numbers for equality.



We'll also meet 5 new built-in R functions:

* `all()`

* `any()`

* `which()`

* `allequal()`

* `isTRUE()`



All right! Let's get started by defining the concept of a logical value.





# Section 1: Definition

> **Main Idea:** *We can define the concept of a logical value*

In this section, we'll define the concept of a logical value.



Logical data is very simple -- it consists of two values, `TRUE` and `FALSE`.

Remember that R is case-sensitive, so you have to type these values in exactly the same way, and in particular you have to use all upper-case letters.

For instance, here's the value `TRUE`:
```{r}
TRUE
```

But now look what happens when we don't use caps:
```{r error = TRUE}
True
```
R doesn't know what this `True` object is, even though it is fine with the `TRUE` object.



There is another way to obtain `TRUE` and `FALSE` values.

R has built-in variables named `T` and `F`, and these will evaluate to `TRUE` and `FALSE`, respectively:
```{r}
T
```


```{r}
F
```

The problem with `T` and `F` is that they are variables, and thus can be assigned a new value.

For instance, this is perfectly legitimate:
```{r}
T <- FALSE

T
```
Thus, if you use `T` and `F` in your code, it's possible to have correctly working code that breaks if somehow these variables are destructively modified.

Personally, I don't like to use `T` and `F`, and I think the only appeal they have is that they save a few keystrokes.

I think the best thing is to write out `TRUE` and `FALSE`, and your code will be more robust and readable.



So that's the definition of logical values.

Now let's see how to operate on logical values.





# Section 2: Operations on Logical Data

> **Main Idea:** *We can operate on logical data*

In this section, we'll define three important logical operators, and see how to use them in a vectorized approach.



Just as there is a wide range of operations for working with numeric data, we can also perform operations on data in the logical class as well.

## Logical negation

There is one unary logical operator, called "logical negation" and denoted by `!`.

This operation just inverts logical values: if the input argument is `TRUE`, then the output return value is `FALSE`:
```{r}
! TRUE
```

Similarly, if the input argument is `FALSE`, then the output return value is `TRUE`:
```{r}
! FALSE
```

We can even have a double negative:
```{r}
! ! TRUE
```




## The logical conjunction operator `&`

Just as we had binary arithmetic operations for numeric data, we also have binary operations for logical data.

One such binary logical operation has the fancy name "logical conjunction", and this returns the value `TRUE` if both of the terms are themselves `TRUE`, and `FALSE` if at least one of the terms is `FALSE`:
```{r}
TRUE & TRUE
```

```{r}
TRUE & FALSE
```

```{r}
FALSE & TRUE
```

```{r}
FALSE & FALSE
```



A nice way to think of the logical conjunction operator `&` is that it's really like the word "AND": the R expression `x & y` evaluates to `TRUE` if and only if both `x` AND `y` evaluate to `TRUE`.

In our course, I will use the expression "logical conjunction" when I'm talking about the operator `&`, because that's the technical jargon.

But in practice when we're looking at actual code I will usually say something like "`x` AND `y`".



## The logical disjunction operator `|`

Along with the logical conjunction operator `&` there is another binary logical operation which has the fancy name "logical disjunction".

The logical disjunction operator returns the value `TRUE` if at least one of the terms is `TRUE`, and `FALSE` if both terms are `FALSE`:
```{r}
TRUE | TRUE
```

```{r}
TRUE | FALSE
```


```{r}
FALSE | TRUE
```


```{r}
FALSE | FALSE
```

A nice way to think of the logical disjunction operator `|` is that it's really like the word "OR": the R expression `x | y` evaluates to `TRUE` if and only if either `x` OR `y` evaluate to `TRUE`.

In this course, I will use the expression "logical disjunction" when I'm talking about the operator `|`, because that's the technical jargon.

But in practice when we're looking at actual code I will usually say something like "`x` OR `y`".



## Vectorized logical operations

We can use vectorized operations on logical values.

Let's create two logical vectors to experiment with:
```{r}
first.logical.vector <- 
    c(TRUE, TRUE, FALSE, TRUE, FALSE)

second.logical.vector <- 
    c(TRUE, FALSE, TRUE, TRUE, FALSE)
```

If we perform a logical operation on these vectors using the logical `and` operator denoted by `&`, the result will be a vector where each element is calculated by applying the `&` operator to corresponding pairs of elements of `first.logical.vector` and `second.logical.vector`:
```{r}
first.logical.vector & second.logical.vector
```

Make sure you understand how these values are determined.



So that's how to operate on logical data.

Now let's see how to generate logical vectors from other vectors.




## Exercise 1.1: Logical operations

Perform a vectorized operation on the two logical vectors `first.logical.vector` and `second.logical.vector` by using the logical `or` operator denoted by `|`.

Can you predict what the resulting vector will be before you run the code?

**Solution**










# Section 3: Constructing Logical Vectors

> **Main Idea:** *We can construct logical vectors**

In this section, we'll learn how to generate logical vectors from other types of vectors, in particular numeric vectors.



We can construct logical vectors from other vectors by using comparison operations.

This is a very powerful technique that we will exploit throughout the course.

Let's define a new vector `numeric.vector`:
```{r}
numeric.vector <- c(4, 2, -3, 7, 1)
```


Now consider this code:
```{r eval = FALSE}
numeric.vector < 0
```
What does this code do?

Try to predict the output before you execute the code.

Let's think this through.

First, R will perform a vectorized operation, in which the comparison is performed element-wise:

* The first element of `numeric.vector` has the value 4, and this is strictly greater than 0, so the comparison evaluates to `FALSE`.

* The second element of `numeric.vector` has the value 2, and this is strictly greater than 0, so the comparison evaluates to `FALSE`.

* The third element of `numeric.vector` has the value -3, and this is strictly less than 0, so the comparison evaluates to `TRUE`.

* The fourth element of `numeric.vector` has the value 7, and this is strictly greater than 0, so the comparison evaluates to `FALSE`.

* The fifth element of `numeric.vector` has the value 1, and this is strictly greater than 0, so the comparison evaluates to `FALSE`.

OK, now let's see what the code actually does:
```{r}
numeric.vector < 0
```
Just what we predicted!


We can use this technique with any numerical comparison.


For instance, we can test each element to see if it is greater than or equal to 2:
```{r}
numeric.vector >= 2
```


when you want to compare 2 R objects for equality, you have to use a *double* equal sign, **not** a single equal sign.

That is, when you compare 2 R objects for equality, you have to use the characters "==", **not** a single equals sign "=".

Let's see how this works.

We'll initialize a variable *x*, and then test it for equality with the number 6:
```{r}
x <- 6

x == 6
```



Do you see that I used the double equals sign "==", not the single equals sign "="?

R then performed a test for equality between the value of the variable *x* and the number 6, and since the variable *x* does indeed have the value 6 the result is the logical value `TRUE`. 

Let's try this again:
```{r}
x == 8
```

This time, the result is `FALSE`, because the variable *x* does not have the value 8.

The important point to remember is that in both cases we used the double equals sign "==".

Why can't we use a single equals sign?

R interprets a single equals sign as though it's an assignment operator.

So we can initialize a variable like this:
```{r}
y = 2
```

Now let's look at this variable:
```{r}
y
```

So a single equals sign "=" is equivalent to the assignment operator "<-".

Notice what happens if we try to use a single equals sign to test two R objects for equality:
```{r}
y = 4
```

Hmmm . . . it appears that *nothing* happened.

But actually, something did happen: we destructively modified the value of *y*:
```{r}
y
```

So *y* is now equal to 4.

Thus, the single equals sign "=" worked like the assignment operator "<-".

For an extreme example of this, consider testing two numerical constants for equality:
```{r}
6 == 6
```

Here's another example:
```{r}
6 == 7
```

So this works properly.

But now look what happens if we try to compare two numbers using a single equals sign:
```{r error = TRUE}
6 = 6
```

We get an error!

That's because a single equals sign operates as an assignment operator, and we have to have a variable on the left-hand side of an assignment.

But 6 isn't a variable -- it's a numeric constant, and it doesn't make sense to try to assign a value to a number.

The important thing to remember is that you have to use a double equals sign to test for equality.

To test for inequality, use the `!=` operator:
```{r}
6 != 7
```



We can even test for equality with character vectors.

```{r}
test.character.vector <-
    c( "To", "be", "or", "not", "to", "be" )
```


Now we can test each element for equality with the character string "be":
```{r}
test.character.vector == "be"
```



So that's how to construct logical vectors out of other vectors.

Now let's meet three important logical functions.




## Exercise 1.2: Constructing a logical vector

Let's create a numeric vector:
```{r}
second.numeric.vector <-
    c( 6, 3, 1, 9, -2, 0, -7, 5, 8, -1)
```

Construct a logical vector which has the value `TRUE` in the locations where the corresponding element of `second.numeric.vector` is negative.

**Solution**






# Section 4: Logical Functions

> **Main Idea:** *We'll meet three logical functions**

In this section, we'll meet three important logical functions.



There are three functions that operate on logical vectors alone.

* The `any()` function.

* The `all()` function.

* The `which()` function

Let's start by creating two logical test vectors:
```{r}
first.logical.vector <-
    c(TRUE, TRUE, FALSE, TRUE, FALSE)

second.logical.vector <-
    c(TRUE, FALSE, TRUE, TRUE, FALSE)
```



## `any()`

The `any()` function takes a logical vector and returns the value `TRUE` if at least one of the elements of the vector is `TRUE`, and `FALSE` otherwise.
```{r}
any( first.logical.vector )
```



## `all()`

The second function for logical vectors is `all()`, and this acts in a way very similar to `any()`: it takes a logical vector and returns the value `TRUE` if every element of the vector is `TRUE`, and `FALSE` otherwise.
```{r}
all( second.logical.vector )
```

You should be clear about the similarities and differences between the `any()` and `all()` functions:

* Both functions take a logical vector as their input argument and both functions return a logical value as the output.

* The function `any()` returns the value `TRUE` if at least one element of the input vector is `TRUE`. In other words, the function `any()` returns `TRUE` if and only if ANY of the elements in the input vector are `TRUE`.

* The function `all()` returns the value `TRUE` if every element of the input vector is `TRUE`. In other words, the function `all()` returns `TRUE` if and only if ALL of the elements in the input vector are `TRUE`.



## `which()`

The third function for logical vectors is `which()`, and this will be very useful for us.

This function takes a logical vector and returns a numeric vector consisting of the indices of the elements of the input vector that are `TRUE`.

Let's see how this works:
```{r}
which( first.logical.vector )
```

Do you see how the `which()` function produced this result?

There are three elements in `first.logical.vector` that have the value `TRUE`, and they are located in positions 1, 2, and 4.

Thus, the `which()` function returns a vector consisting of these values.



So those are three important logical functions.

Now let's explore some technical issues when comparing two numeric values for equality.




## Exercise 1.3: Logical functions

Here's `second.logical.vector`:
```{r}
second.logical.vector
```

What is the result of calling the `all()` function with this vector?

What is the result of calling the `any()` function with this vector?

What is the result of calling the `which()` function with this vector?

Try to figure these out yourself before trying it in R.

**Solution**




# Section 5: Testing floating-point numbers for equality

> **Main Idea:** *We have to be careful when comparing floating point numbers for equality*.

In this section, we'll discuss some technical issues that arise when comparing floating point numbers.



We can encounter problems if we try to test floating-point numbers for exact equality.

Recall that a "floating point" number is just something with a decimal point.

Let's calculate the square root of 5, and display it directly:
```{r}
square.root.of.5 <- sqrt( 5 )

square.root.of.5
```

Let's check this by squaring this variable:
```{r}
square.root.of.5^2
```

Everything seems reasonable -- we calculated the square root of 5, and then just to be sure we squared this value and found that it was equal to 5.

But now look at what happens when we test for equality:
```{r}
square.root.of.5^2 == 5
```
WHAT?!?!??!?

How can this be `FALSE`?

The answer is that the square root of 5 does not have a finite decimal expansion -- in reality, the square root of 5 requires an infinite number of decimals.

However, our computer system can only store a finite number of decimals.

When we calculate the square root of 5, the series is *truncated*, which means that it's cut off at a certain point.

The value that we think is the square root of 5 isn't *exactly* the same as the square root of 5, but rather is just the closest approximation that our computer can represent.

Thus, when we square this value, it's not *exactly* equal to 5, but just very close.

We can see this by taking the difference of our squared value with 5:
```{r}
square.root.of.5^2 - 5
```

This notation indicates that the difference between the squared value and 5 is 8.9 times 10 raised to the negative sixteenth power.

If we wrote this out in conventional floating point notation, we would have a zero and a decimal point followed by 15 zeroes before we see the leading digit of 8.
```{r}
formatC(
  square.root.of.5^2 - 5,
  format = "f",
  digits = 18
)
```


So the difference is very small, so small that in practice it's essentially negligible.

But the difference is not 0, hence the two numbers are not identical, and that's why the test for equality returns `FALSE`.

In general, we have to be very cautious about testing two floating point numbers for equality, especially if we perform an operation such as taking a square root.



If we do want to test two floating point numbers for equality, then we can use the `all.equal()` function, which takes two numbers and returns `TRUE` if they are sufficiently close:
```{r}
all.equal( square.root.of.5^2, 5)
```

What precisely does "sufficiently close" mean?

In practice, `all.equal()` will return `TRUE` if the two numbers are within about
$1.5 \times 10^{-8}$ of each other, and you can adjust this tolerance level.

However, there's a catch -- if the two numbers are sufficiently close, `all.equal()` will return the logical value `TRUE`, but if they are not sufficiently close the function will return a string declaring the difference:
```{r}
all.equal( 4.9, 5.0 )
```
This is a very unusual function, because it returns objects of different classes.

In other words, if the two numbers are sufficiently close, then `all.equal()` will return the logical value `TRUE`, but if they are not sufficiently close then it will return a character string.

What we would really like is to have some way to test two floating point numbers, and have R return `TRUE` if the numbers are sufficiently close and `FALSE` if they are not.

To do this, we can use the function `isTRUE()`, which takes a single input argument and returns the value `TRUE` if the input argument is the logical value `TRUE` and returns the value `FALSE` otherwise.

If two numbers are sufficiently close, then `all.equal()` will return `TRUE`, and since this is the logical value `TRUE` then `isTRUE()` will return `TRUE`:
```{r}
isTRUE( all.equal( square.root.of.5^2, 5 ) )
```

On the other hand, if the two values are not sufficiently close, then `all.equal()` will return a character string value, and since this is not the logical value `TRUE` then `isTRUE()` will return `FALSE`:
```{r}
isTRUE( all.equal( 4.9, 5.0 ) )
```

And this is exactly what we want!

In practice, you should think twice before comparing two floating point numbers for exact equality.



So that's how we compare two floating point numbers for equality.

Now let's review what we've learned in this module.






# Module Review

In this module, we studied logical values.

* In section 1, we defined the concept of a logical value.

* In section 2, we defined three important logical operators, and saw how to use them in a vectorized approach.

* In section 3, we learned how to generate logical vectors from other types of vectors, in particular numeric vectors.

* In section 4, we met three important logical functions.

* In section 5, we discussed some technical issues that arise when comparing floating point numbers.


Now that you've completed this module, you should be able to:

* Explain the concept of logical data.

* Perform logical operations on logical values and vectors.

* Construct logical vectors from other types of vectors, especially numeric vectors.

* Explain how the logical functions `all()`, `any()`, and `which()` operate.

* Explain how to compare floating point numbers for equality.


We met 5 new built-in R functions:

* `all()`

* `any()`

* `which()`

* `allequal()`

* `isTRUE()`



All right! That's it for Module 1: Logical Values.

Now let's move on to Module 2: Sample Counts and Proportions.






# Solutions to the Exercises

## Exercise 1.1: Logical operations

Perform a vectorized operation on the two logical vectors `first.logical.vector` and `second.logical.vector` by using the logical `or` operator denoted by `|`.

Can you predict what the resulting vector will be before you run the code?

**Solution**

```{r}
first.logical.vector | second.logical.vector
```



## Exercise 1.2: Constructing a logical vector

Let's create a numeric vector:
```{r}
second.numeric.vector <-
    c( 6, 3, 1, 9, -2, 0, -7, 5, 8, -1)
```

Construct a logical vector which has the value `TRUE` in the locations where the corresponding element of `second.numeric.vector` is negative.

**Solution**

```{r}
second.numeric.vector < 0
```



## Exercise 1.3: Logical functions

Here's `second.logical.vector`:
```{r}
second.logical.vector
```

What is the result of calling the `all()` function with this vector?

What is the result of calling the `any()` function with this vector?

What is the result of calling the `which()` function with this vector?

Try to figure these out yourself before trying it in R:

**Solution**

Since at least one value in `second.logical.vector` is `FALSE`, the `all()` function returns `FALSE`:
```{r}
all( second.logical.vector )
```


Since at least one value in `second.logical.vector` is `TRUE`, the `all()` function returns `TRUE`:

```{r}
any( second.logical.vector )
```

Since `second.logical.vector` has the value `TRUE` in locations 1, 3, and 4, the `which()` function returns the values 1, 3, and 4:
```{r}
which( second.logical.vector)
```





































