---
title: "Week 2 Module 4: Your Final Grade EXERCISES"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---


Let's clear the global computing environment:
```{r}
rm( list = ls() )
```




# Exercise for Week 2 Module 4: Your Final Grade


## Exercise 1: Your Final Grade

### Part (a)

Bob is registered for graduate credit, and obtained these raw scores:

* He scored 52 total points on his problem sets.

* He scored a 72 on the midterm.

* He scored a 74 on the comprehensive course assessment.

Calculate Bob's final letter grade for the course.

**Answer**



### Part (b)

Ashley is registered for undergraduate credit, but obtains the exact same scores as Bob does.

Calculate Ashley's final letter grade.


**Answer**





# Solution to the Exercise

## Exercise 1: Your Final Grade

### Part (a)

Bob is registered for graduate credit, and obtained these raw scores:

* He scored 52 total points on his problem sets.

* He scored a 72 on the midterm.

* He scored a 74 on the comprehensive course assessment.

Calculate Bob's final letter grade for the course.


**Answer**

Let's define some variables to store Bob's raw scores:
```{r}
problem.set.raw.score <- 52

midterm.exam.raw.score <- 72

comprehensive.assessment.raw.score <- 74
```


First, we'll standardize Bob's scores.

Bob's standardized score for the problem set points is:
```{r}
problem.set.standardized.score <-
    problem.set.raw.score / 68 * 100
    
cat(
    "Standardized problem set score:",
    formatC(
        problem.set.standardized.score,
        format = "f",
        digits = 2
    )
)
```


Bob's standardized score for the midterm exam is:
```{r}
midterm.exam.standardized.score <-
    midterm.exam.raw.score / 80 * 100
    
cat(
    "Standardized midterm exam score:",
    formatC(
        midterm.exam.standardized.score,
        format = "f",
        digits = 2
    )
)
```


Bob's standardized score for the comprehensive course assessment is:
```{r}
comprehensive.assessment.standardized.score <-
    comprehensive.assessment.raw.score / 80 * 100
    
cat(
    "Standardized comprehensive.assessment score:",
    formatC(
        comprehensive.assessment.standardized.score,
        format = "f",
        digits = 2
    )
)
```


Now we'll calculate Bob's preliminary score 1:
```{r}
preliminary.score.1 <-
    (0.20 * problem.set.standardized.score) +
    (0.30 * midterm.exam.standardized.score) +
    (0.50 * comprehensive.assessment.standardized.score)

cat(
    "Preliminary score 1:",
    formatC(
        preliminary.score.1,
        format = "f",
        digits = 2
    )
)
```


Now we'll calculate Bob's preliminary score 2:
```{r}
preliminary.score.2 <-
    (0.35 * midterm.exam.standardized.score) +
    (0.65 * comprehensive.assessment.standardized.score)

cat(
    "Preliminary score 2:",
    formatC(
        preliminary.score.2,
        format = "f",
        digits = 2
    )
)
```

Since the value of preliminary score 2 is greater than the value of preliminary score 1, I use the preliminary score 2 value as the graduate final course score.

```{r}
graduate.final.course.score <-
    max( preliminary.score.1, preliminary.score.2 )

cat(
    "Final graduate course score for Bob:",
    formatC(
        graduate.final.course.score,
        format = "f",
        digits = 2
    )
)
```


Let's summarize Bob's scores:

|                             | Value |
|:----------------------------|:-----:|
| Preliminary Score 1         | 88.54 |
| Preliminary Score 2         | 91.62 |
|:----------------------------|:-----:|
| Final Graduate Course Score | 91.62 |

When I look up the final graduate course score of 91.62 in the letter grade table, I find that this corresponds to a grade of "A-".

So Bob gets an "A-" for the course as his final grade.


### Part (b)

Ashley is registered for undergraduate credit, but obtains the exact same scores as Bob does.

Calculate Ashley's final letter grade.

**Answer**

Since Ashley is registered for undergraduate credit, her pro-rated undergraduate final course score is:
```{r}
undergraduate.final.course.score <-
    4/3 * graduate.final.course.score

cat(
    "Final course score for Ashley:",
    formatC(
        undergraduate.final.course.score,
        format = "f",
        digits = 2
    )
)
```

So in this case Ashley gets a final letter grade of "A".








