---
title: 'Lecture 8, Module 1: Character Strings'
subtitle: Exercises
output:
  html_document:
    df_print: paged
---

```{r}
rm( list = ls() )
```


```{r}
load( "Lecture 8 Module 1 R Objects.Rdata" )
```

# Exercises

## Exercise 1: Stripping out an internal code

The vector `exercise.1.data` consists of character strings, each of which begins with an internal code of fixed length, followed by an identifier of variable length consisting entirely of digits.

For each character string value in `exercise.1.data`, strip out the internal code of fixed length, and return a character string vector consisting of just the variable length identifiers.

You'll have to figure out how long the internal code by examining the data yourself.

**Solution**







## Exercise 2: Ted Williams' On-Base Percentage

Here are Ted Williams's career statistics:

| Statistics        | Value |
|:------------------|------:|
| Plate appearances | 9,792 |
| At-bats           | 7,706 |
| Hits              | 2,654 |
| Doubles           |   525 |
| Triples           |    71 |
| Home Runs         |   521 |
| Bases on balls    | 2,021 |
| Hit by a pitch    |    39 |
| Sacrifice flies   |    20 |

Report Ted Williams' On-Base Percentage (OBP) using a `cat()` statement, with the value displayed using the standard baseball conventions.

**Solution**

## Exercise 3: Substring pattern matching with the `grep()` function

Consider this set of data:

```{r}
exercise.3.data <-
  c(
    "healthy kale",
    "Sugar Bms",
    "Sugar Bom",
    "Krispy Ym!",
    "kale Tfuo",
    "Sug Bz",
    "Crispy Yum",
    "healthy kale and tofu"
  )
```

### Part (a): Counting rows with "kale"

How many rows in `exercise.3.data` contain the string "kale"?

**Solution**

### Part (b): Converting to "kale" or "no kale"

Create a new vector consisting of the values "kale" or "no kale", where this new vector has the value "kale" precisely when the input data vector `exercise.3.data` has an element with a substring that matches the pattern string "kale".

**Solution**

### Part (c): Converting all "kale" entries to "HKT"

Create a new vector consisting of the values of the original data vector `exercise.3.data`, but with all elements that match the pattern "kale" replaced with "HKT".

**Solution**




# Solution to the Exercises


## Exercise 1: Stripping out an internal code

The vector `exercise.1.data` consists of character strings, each of which begins with an internal code of fixed length, followed by an identifier of variable length consisting entirely of digits.

For each character string value in `exercise.1.data`, strip out the internal code of fixed length, and return a character string vector consisting of just the variable length identifiers.

You'll have to figure out how long the internal code by examining the data yourself.

**Solution**

Let's first take a look at the values in `exercise.1.data` by directly displaying the vector:

```{r}
exercise.1.data
```

You can see that the first 7 characters of each string are always the same, so this is the internal code.

Then we can extract the identifier string by using the `substr()` function and the `nchar()` function:

```{r}
identifier.string.vector <-
  substr(
    x = exercise.1.data,
    start = 8,
    stop = nchar( exercise.1.data )
  )
```

Now let's display this directly:

```{r}
identifier.string.vector
```




## Exercise 2: Ted Williams' On-Base Percentage

Here are Ted Williams's career statistics:

| Statistics        | Value |
|:------------------|------:|
| Plate appearances | 9,792 |
| At-bats           | 7,706 |
| Hits              | 2,654 |
| Doubles           |   525 |
| Triples           |    71 |
| Home Runs         |   521 |
| Bases on balls    | 2,021 |
| Hit by a pitch    |    39 |
| Sacrifice flies   |    20 |

Report Ted Williams' On-Base Percentage (OBP) using a `cat()` statement, with the value displayed using the standard baseball conventions.

**Solution**

Let's start by creating a set of variables to hold Williams' career batting statistics:

```{r}
ted.williams.at.bats <- 7706

ted.williams.hits <- 2654

ted.williams.doubles <- 525

ted.williams.triples <- 71

ted.williams.home.runs <- 521

ted.williams.bases.on.balls <- 2021

ted.williams.hit.by.a.pitch <- 39

ted.williams.sacrifice.flies <- 20
```

Now we can calculate Ted Williams' career on-base percentage:

```{r}
ted.williams.on.base.percentage <-
    (ted.williams.hits +
         ted.williams.bases.on.balls +
         ted.williams.hit.by.a.pitch) /
    (ted.williams.at.bats +
         ted.williams.bases.on.balls +
         ted.williams.hit.by.a.pitch +
         ted.williams.sacrifice.flies)
```

Next, we'll format this result using the `formatC()` function so that it has exactly 3 digits trailing the decimal point, and then directly display it to check our work:

```{r}
ted.williams.on.base.percentage.string <-
  formatC(
    x = ted.williams.on.base.percentage,
    format = "f",
    digits = 3
  )

ted.williams.on.base.percentage.string
```

Now let's use the `substr()` function to remove the leading 0, and then directly display this string to check our work:

```{r}
ted.williams.on.base.percentage.formatted.string <-
  substr(
    x = ted.williams.on.base.percentage.string,
    start = 2,
    stop = 5
  )

ted.williams.on.base.percentage.formatted.string
```

Finally, we can report this with a `cat()` statement:

```{r}
cat(
  "Ted Williams on-base percentage:",
  ted.williams.on.base.percentage.formatted.string
)
```






## Exercise 3: Substring pattern matching with the grep() function

Consider this set of data:
```{r}
exercise.3.data <-
  c(
    "healthy kale",
    "Sugar Bms",
    "Sugar Bom",
    "Krispy Ym!",
    "kale Tfuo",
    "Sug Bz",
    "Crispy Yum",
    "healthy kale and tofu"
  )
```

### Part (a): Counting rows with "kale"

How many rows in `exercise.3.data` contain the string "kale"?

**Solution**

```{r}
kale.match.index.vector <-
  grep(
    pattern = "kale",
    x = exercise.3.data
  )

length( kale.match.index.vector )
```



### Part (b): Converting to "kale" or "no kale"

Create a new vector consisting of the values "kale" or "no kale", where this new vector has the value "kale" precisely when the input data vector `exercise.3.data` has an element with a substring that matches the pattern string "kale".

**Solution**

```{r}
kale.match.logical.indexing.vector <-
  grepl(
    pattern = "kale",
    x = exercise.3.data
  )

ifelse(
  test = kale.match.logical.indexing.vector,
  yes = "kale",
  no = "no kale"
)
```




### Part (c): Converting all "kale" entries to "HKT"

Create a new vector consisting of the values of the original data vector `exercise.3.data`, but with all elements that match the pattern "kale" replaced with "HKT".

**Solution**

```{r}
kale.match.logical.indexing.vector <-
  grepl(
    pattern = "kale",
    x = exercise.3.data
  )

ifelse(
  test = kale.match.logical.indexing.vector,
  yes = "HKT",
  no = exercise.3.data
)
```








