---
title: 'CSCI 5a Lecture 4 Module 4: Loan Amortization'
subtitle: "Exercises"
output:
  pdf_document: default
  html_document:
    df_print: paged
---



# Exercises

## Exercise 4.1: Annual Payment Amount

Ashley takes out a loan for \$10,000 and pays it back in 3 annual installments, the first one occurring one year after the loan is issued.

The annual interest rate is 4\%.

Save the loan amount and the annual interest rate in variables, but call the variables something different from `loan.amount` and `annual.interest.rate`, because we're using those variables for our text example.

Calculate the annual payment amount.

Save this annual payment amount in a variable, but call the variable something different from `annual.payment.amount`, because we're using that for our text example.

**Solution**




## Exercise 4.2: Amortizing a loan

Ashley takes out a loan for \$10,000 and pays it back in 3 annual installments, the first one occurring one year after the loan is issued.

The interest rate is 4\%.

Calculate the amortization schedule.

**Solution**





# Solutions to the Exercises

## Exercise 4.1: Annual Payment Amount

Ashley takes out a loan for \$10,000 and pays it back in 3 annual installments, the first one occurring one year after the loan is issued.

The annual interest rate is 4\%.

Save the loan amount and the annual interest rate in variable, but call the variables something different from `loan.amount` and `annual.interest.rate`, because we're using those variables for our text example.

Calculate the annual payment amount.

Save this annual payment amount in a variable, but call the variable something different from `annual.payment.amount`, because we're using that for our text example.

**Solution**

First, we can define some variables:
```{r}
ashley.loan.amount <- 10000

ashley.number.of.payments <- 3

ashley.annual.interest.rate <- 0.04
```


Now we can calculate the discount factor:
```{r}
ashley.discount.factor <- 
    1 / (1 + ashley.annual.interest.rate)

cat(
    "Discount factor:",
    formatC(
        ashley.discount.factor,
        format = "f",
        digits = 6
    )
)
```


Now we can calculate the annual payment amount:
```{r}
ashley.annual.payment.amount <-
    ashley.loan.amount * 
    ashley.annual.interest.rate /
    (1 - ashley.discount.factor^ashley.number.of.payments)

cat(
    "Annual payment amount:",
    formatC(
        ashley.annual.payment.amount,
        format = "f",
        big.mark = ",",
        digits = 2
    )
)
```



## Exercise 4.2: Amortizing a loan

Ashley takes out a loan for \$10,000 and pays it back in 3 annual installments, the first one occurring one year after the loan is issued.

The interest rate is 4\%.

You've already calculated the annual payment amount for Ashley in Exercise 3.1.

Now construct an amortization schedule to show that this annual payment amount will indeed pay off her initial loan in 3 years.


**Solution**

At the start of the first year, Ashley has a loan for \$10,000, with an interest rate of 4.00\%.
```{r}
ashley.loan.balance <- 1000
```

At the end of the year, the loan balance has accrued 4\% interest.

Let's update the loan balance variable:
```{r}
ashley.loan.balance <-
    ashley.loan.balance + 
    (ashley.annual.interest.rate * ashley.loan.balance)

cat(
    "Loan balance at end of year 1, before payment:",
    formatC(
        ashley.loan.balance,
        format = "f",
        digits = 2
    )
)
```

This represents the loan balance at the end of the year *before* Ashley makes the first annual payment.

When Ashley makes the annual payment at the end of year 1, the loan balance decreases by the amount of the payment:
```{r}
ashley.loan.balance <-
    ashley.loan.balance - ashley.annual.payment.amount

cat(
    "Loan balance at end of year 1, after payment:",
    formatC(
        ashley.loan.balance,
        format = "f",
        digits = 2
    )
)
```

The loan balance at the end of year 1 after the payment is made now becomes the loan balance at the beginning of year 2:
```{r}
cat(
    "Loan balance at beginning of year 2:",
    formatC(
        ashley.loan.balance,
        format = "f",
        digits = 2
    )
)
```

This loan balance accrues 4\% interest over the duration of the second year.

The loan balance at the end of the second year, before Ashley makes the annual payment, is:
```{r}
ashley.loan.balance <-
    ashley.loan.balance +
    (ashley.loan.balance * ashley.annual.interest.rate)

cat(
    "Loan balance at end of year 2, before payment:",
    formatC(
        ashley.loan.balance,
        format = "f",
        digits = 2
    )
)
```

When Ashley makes the annual payment at the end of the second year, the loan balance decreases by the amount of the payment:
```{r}
ashley.loan.balance <-
    ashley.loan.balance - ashley.annual.payment.amount

cat(
    "Loan balance at end of year 2, after payment:",
    formatC(
        ashley.loan.balance,
        format = "f",
        digits = 2
    )
)
```

The loan balance at the end of year 2 after the payment is made now becomes the loan balance at the beginning of year 3:
```{r}
cat(
    "Loan balance at beginning of year 3:",
    formatC(
        ashley.loan.balance,
        format = "f",
        digits = 2
    )
)
```

The loan balance at the end of the third year, before Ashley makes the annual payment, is:
```{r}
ashley.loan.balance <-
    ashley.loan.balance +
    (ashley.loan.balance * ashley.annual.interest.rate)

cat(
    "Loan balance at end of year 3, before payment:",
    formatC(
        ashley.loan.balance,
        format = "f",
        digits = 2
    )
)
```


When Ashley makes the annual payment at the end of the third year, the loan balance is 0.00 and the loan is paid off:
```{r}
ashley.loan.balance <-
    ashley.loan.balance - ashley.annual.payment.amount

cat(
    "Loan balance at end of year 3, after payment:",
    formatC(
        ashley.loan.balance,
        format = "f",
        digits = 2
    )
)
```







