---
title: 'Lecture 3 Module 1: Vector Basics'
subtitle: Exercises
output:
  pdf_document: default
  html_notebook: default
---


# Exercises

Let's load in the R objects for this module:
```{r}
load( "Module 1 R Objects.Rdata")
```



## Exercise 1.1: Exploring a vector

Display the first 8 elements of the `test.vector` that we loaded in from the R object file.

Then determine the length of the vector, as well as the sum and product of the elements of the vector.

**Solution**



## Exercise 1.2: Mean, variance, and standard deviation

Calculate the sample mean of the values in `test.vector`. Report your result using a `cat()` statement, displaying this value with 2 decimal places.

Then calculate the sample variance of the values in `test.vector`. Report your result using a `cat()` statement, displaying this value with 2 decimal places.

Finally, calculate the standard deviation of the values in `test.vector`. Report your result using a `cat()` statement, displaying this value with 2 decimal places.

**Solution**





## Exercise 1.3: Report the `rivers` vector

Report the first 20 items of the built-in `rivers` vector, displaying these values using 2 decimal places.




# Solutions to the Exercises


## Exercise 1.1: Exploring a vector

Display the first 8 elements of the `test.vector` that we loaded in from the R object file.

Then determine the length of the vector, as well as the sum and product of the elements of the vector.

**Solution**

Let's display the first 8 elements of `test.vector`:
```{r}
head( x = test.vector, n = 8 )
```

Next, the sum of the elements of `test.vector` is:
```{r}
sum( test.vector )
```

The product of the elements of `test.vector` is:
```{r}
prod( test.vector )
```




## Exercise 1.2: Mean, variance, and standard deviation

Calculate the sample mean of the values in `test.vector`. Report your result using a `cat()` statement, displaying this value with 2 decimal places.

Then calculate the sample variance of the values in `test.vector`. Report your result using a `cat()` statement, displaying this value with 2 decimal places.

Finally, calculate the standard deviation of the values in `test.vector`. Report your result using a `cat()` statement, displaying this value with 2 decimal places.

**Solution**

First, let's do the sample mean:
```{r}
sample.mean <-
    mean( test.vector )

cat(
    "Sample mean:",
    formatC(
        sample.mean,
        format = "f",
        digits = 2
    )
)
```


Next, let's do the sample variance:
```{r}
sample.variance <-
    var( test.vector )

cat(
    "Sample variance:",
    formatC(
        sample.variance,
        format = "f",
        digits = 2
    )
)
```



Finally, here's the sample standard deviation:
```{r}
sample.standard.deviation <-
    sd( test.vector )

cat(
    "Sample standard deviation:",
    formatC(
        sample.standard.deviation,
        format = "f",
        digits = 2
    )
)
```





## Exercise 1.3: Report the `rivers` vector

Report the first 20 items of the built-in `rivers` vector, displaying these values using 2 decimal places.

**Solution**

```{r}
cat(
    "rivers:",
    formatC(
        head( rivers, 20 ),
        format = "f",
        digits = 2
    ),
    fill = TRUE
)
```

















































