---
title: 'Week 2 Module 6: Pie Charts EXERCISES'
subtitle: "CSCI E-5a: Programming in R"
output:
  pdf_document: default
  html_document:
    df_print: paged
---



# Exercises for Week 2 Module 6: Pie Charts


## Exercise 5.1: Babe Ruth At-Bats

Recall that an at-bat is either an out or a hit, and there are four different kinds of hits.

Thus, there are five possible outcomes for an at-bat:

* An out

* A single

* A double

* A triple

Let's make a simple pie chart to display the relative proportions of these outcomes for Babe Ruth.

Recall that Babe Ruth had these batting statistics:

| Statistics        |  Value |
|:------------------|-------:|
| Plate appearances | 10,626 |
| At-bats           |  8,399 |
| Hits              |  2,873 |
| Doubles           |    506 |
| Triples           |    136 |
| Home runs         |    714 |
| Bases on balls    |  2,062 |
| Hit by a pitch    |     43 |
| Sacrifice flies   |      0 |

Remember these two derived statistics:

* To calculate the number of hits, subtract the number of hits from the number of at-bats

* To calculate the number of singles, subtract the number of doubles, triples, and home runs from the number of hits.

Here are some variables for you:
```{r}
babe.ruth.plate.appearances <- 10626

babe.ruth.at.bats <- 8399

babe.ruth.hits <- 2873
babe.ruth.doubles <- 506
babe.ruth.triples <- 136
babe.ruth.home.runs <- 714
```

**Solution**





## Exercise 5.2: Modifying the Babe Ruth pie chart

Now add labels and a main title to the Babe Ruth pie chart, and set the rotation to clockwise.

**Solution**





## Exercise 5.3: Colors for the Babe Ruth pie chart

Now explicitly specify the pie slice colors for the Babe Ruth pie chart.

**Solution**







# Solutions to the Exercises

## Exercise 5.1: Babe Ruth At-Bats

Recall that an at-bat is either an out or a hit, and there are four different kinds of hits.

Thus, there are five possible outcomes for an at-bat:

* An out

* A single

* A double

* A triple

Let's make a simple pie chart to display the relative proportions of these outcomes for Babe Ruth.

Recall that Babe Ruth had these batting statistics:

| Statistics        |  Value |
|:------------------|-------:|
| Plate appearances | 10,626 |
| At-bats           |  8,399 |
| Hits              |  2,873 |
| Doubles           |    506 |
| Triples           |    136 |
| Home runs         |    714 |
| Bases on balls    |  2,062 |
| Hit by a pitch    |     43 |
| Sacrifice flies   |      0 |

Remember these two derived statistics:

* To calculate the number of hits, subtract the number of hits from the number of at-bats

* To calculate the number of singles, subtract the number of doubles, triples, and home runs from the number of hits.

Here are some variables for you:
```{r}
babe.ruth.plate.appearances <- 10626

babe.ruth.at.bats <- 8399

babe.ruth.hits <- 2873
babe.ruth.doubles <- 506
babe.ruth.triples <- 136
babe.ruth.home.runs <- 714
```

**Solution**

```{r}
babe.ruth.outs <-
  babe.ruth.at.bats - babe.ruth.hits

babe.ruth.singles <-
  babe.ruth.hits -
  (babe.ruth.doubles +
     babe.ruth.triples +
     babe.ruth.home.runs)

pie(
  x =
    c(
      babe.ruth.singles,
      babe.ruth.doubles,
      babe.ruth.triples,
      babe.ruth.home.runs,
      babe.ruth.outs
    )
)
```




## Exercise 5.2: Modifying the Babe Ruth pie chart

Now add labels and a main title to the Babe Ruth pie chart, and set the rotation to clockwise.

**Solution**

```{r}
pie(
  x =
    c(
      babe.ruth.singles,
      babe.ruth.doubles,
      babe.ruth.triples,
      babe.ruth.home.runs,
      babe.ruth.outs
    ),
    
  labels =
        c( "Singles", "Doubles", "Triples",
           "Home Runs", "Outs"),
    
  main = "Babe Ruth Career At-Bats",
    
  clockwise = TRUE
)
```



## Exercise 5.3: Colors for the Babe Ruth pie chart

Now explicitly specify the pie slice colors for the Babe Ruth pie chart.

**Solution**

```{r}
pie(
  x =
    c(
      babe.ruth.singles,
      babe.ruth.doubles,
      babe.ruth.triples,
      babe.ruth.home.runs,
      babe.ruth.outs
    ),
    
  labels =
        c( "Singles", "Doubles", "Triples",
           "Home Runs", "Outs"),
    
  main = "Babe Ruth Career At-Bats",
    
  clockwise = TRUE,
  
  col =
    c( "cyan1", "cadetblue2", "darkslategray3",
       "royalblue", "salmon1" )
)
```








































































