---
title: "Lecture 11 Module 5 -- Two-Way Tables and Barplots"
subtitle: 'CSCI E-5a: Programming in R R'
output:
  pdf_document: default
  html_notebook: default
---


Let's clear the global computing environment:
```{r}
rm( list = ls() )
```



# Module Overview

Hello! And welcome to Module 5: Two-Way Tables and Barplots.

In this module, we'll learn how to construct tables and barplots across two variables.



* In Section 1, we'll learn a variety of methods for constructing a two-way table.

* In Section 2, we'll learn how to construct a two-way barplot.



When you've completed this module, you'll be able to:

* Construct two-way tables

* Construct two-way barplots



There is one new built-in R function in this module:

* `addmargins()`



All right! Let's get started by learning how to construct two-way tables.





# Section 1: Two-Way Tables

**Main Idea:** *We can construct two-way tables*

In this section, we'll learn a variety of methods for constructing a two-way table.



Now let's take a look at a more complicated example.

| Weight | Color | Region |
|:------:|:-----:|:------:|
| 467 | Red | East |
| 483 | Red | East |
| 491 | Red | East |
| 452 | Red | West |
| 441 | Red | West |
| 398 | Red | West |
| 447 | Red | West |
| 338 | Blue | East |
| 325 | Blue | East |
| 304 | Blue | West |
| 313 | Blue | West |
| 317 | Blue | West |
| 322 | Blue | West |
| 309 | Blue | West |

Let's put this into R:
```{r}
weight.vector <-
    c( 467, 483, 491, 452, 441, 398, 447,
       338, 325, 304, 313, 317, 322, 309 )

color.vector <-
    c( "Red", "Red", "Red", "Red", "Red", "Red", "Red",
       "Blue", "Blue", "Blue", "Blue",
       "Blue", "Blue", "Blue" )

region.vector <-
    c( "East", "East", "East",
       "West", "West", "West", "West",
       "East", "East",
       "West", "West", "West", "West", "West" )
```

Notice that in this case we now have two categorical variables: `color.vector` and `region.vector`.

We can create a two-dimensional table of the counts by using the `table` function with the two categorical variables:
```{r}
two.way.count.table <-
    table( color.vector, region.vector )

two.way.count.table
```

You can convert this two-way frequency count table to a two-way relative proportions table using the `proportions()` function:
```{r}
proportions( two.way.count.table )
```

You can add marginal totals to a two-way table by using the `addmargins()` function:
```{r}
addmargins( two.way.count.table )
```




We can also perform two-way summaries with `tapply()`, but in this case we have to bundle the two categorical vectors together using a list:
```{r}
two.way.mean.summary <-
    tapply(
        weight.vector,
        list( color.vector, region.vector ),
        mean
    )

two.way.mean.summary
```

You can also do a two-way analysis like this using the `aggregate()` function:
```{r}
aggregate(
  x = list( weight = weight.vector ),
  by = list( Color = color.vector, Region = region.vector ),
  FUN = mean
)
```



So those are a variety of methods for constructing a two-way table.

Now let's see how to construct two-way barplots.





# Section 2: Two-Way Barplots

**Main Idea:** *We can construct two-way barplots*

In this section, we'll learn how to construct a two-way barplot.



We can make a grouped barplot with this data:
```{r}
barplot(
    two.way.count.table,
    main = "Two-way table of counts",
    xlab = "Region",
    ylab = "Counts",
    beside = TRUE,
    col = c( "skyblue2", "salmon2" )
)

legend(
    x = 1,
    y = 5,
    legend = c( "Blue", "Red" ),
    pch = 15,
    col = c( "skyblue2", "salmon2" ),
    pt.cex = 2
)
```

If you leave out the parameter `beside = TRUE`, the bars are stacked one on top of the other:
```{r}
barplot(
    two.way.count.table,
    ylim = c(0, 10),
    main = "Two-way table of counts",
    xlab = "Region",
    ylab = "Counts",
    col = c( "skyblue2", "salmon2" )
)

legend(
    x = 0.2,
    y = 10,
    legend = c( "Blue", "Red" ),
    pch = 15,
    col = c( "skyblue2", "salmon2" ),
    pt.cex = 2
)
```

Personally, I find these sorts of charts hard to interpret, and I don't think they are used very often in the real world.

We can also make these grouped barplots using the summary from the `tapply()` function:
```{r}
barplot(
    two.way.mean.summary,
    ylim = c(0, 600),
    main = "Grouped barplot of mean weights",
    xlab = "Region",
    ylab = "Mean",
    beside = TRUE,
    col = c( "skyblue2", "salmon2" )
)

legend(
    x = 1,
    y = 600,
    legend = c( "Blue", "Red" ),
    pch = 15,
    col = c( "skyblue2", "salmon2" ),
    pt.cex = 2
)
```


Notice that the bars are grouped by colors first, and then regions.

That's because of how we specified the call to the `tapply()` function.

If we reverse this order, we'll get a different barplot:
```{r}
barplot(
    tapply(
        weight.vector,
        list( region.vector, color.vector ),
        mean
    ),
    ylim = c(0, 600),
    main = "Grouped barplot of mean weights",
    xlab = "Color",
    ylab = "Mean",
    beside = TRUE,
    col = c( "cyan1", "royalblue2" )
)

legend(
    x = 1,
    y = 600,
    legend = c( "East", "West" ),
    pch = 15,
    col = c( "cyan1", "royalblue2" ),
    pt.cex = 2
)
```



So that's how to construct a two-way barplot.

Now let's review what we've learned in this module.



# Module Review


In this module, we learned how to construct tables and barplots across two variables.



* In Section 1, we learned a variety of methods for constructing a two-way table.

* In Section 2, we learned how to construct a two-way barplot.



Now that you've completed this module, you should be able to:

* Construct two-way tables

* Construct two-way barplots



There was one new built-in R function in this module:

* `addmargins()`



All right! That's it for Module 5: Two-Way Tables and Barplots.

In fact, that's the end of the content for Week 11: Summarizing Data.

And it's also the end of the core content for our course!









