---
title: "Week 1 Module 3: Points"
subtitle: "Exercises"
output:
  pdf_document: default
  html_notebook: default
---


Let's clear the global computing environment:
```{r}
rm( list = ls() )
```


# Exercises for Week 1 Module 3: Points


## Exercise 2.1: Graphing a single point

First, create an empty plotting region with no data, where the $x$-values range from 0 to 10, and the $y$-values range from 0 to 8$.

Then use the `points()` function to draw a single point located at $x = 7$ and $y = 3$.

**Solution**





## Exercise 2.2: Annotating with text

Draw a graph with a single point located at $x = 7$ and $y = 3$.

Set the $x$-values for the plotting region to range from 0 to 10, and the $y$-values to range from 0 to 8.

Then annotate this point with its coordinates using the `text()` function.

**Solution**





## Exercise 2.3: Adjusting the size of a point

Create a graph of the single point at $x = 7$ and $y = 12$, and use a character expansion factor of 3.5.

Let the $x$-axis range from 0 to 10, and let the $y$-axis range from 0 to 20.

Remember to include a main title and to properly label the axes.

**Solution**





## Exercise 2.4: Adjusting the color of a point

Create a graph of the single point at $x = 7$ and $y = 12$, using a character expansion factor of 3.5 and a color of "hotpink3".

Let the $x$-axis range from 0 to 10, and let the $y$-axis range from 0 to 20.

Remember to include a main title and to properly label the axes.

**Solution**

```{r}
# Type your answer in here
```




## Exercise 2.5: Adjusting the shape of a point

Create a graph of the single point with a solid square shape at $x = 7$ and $y = 12$, using a character expansion factor of 3.5 and the color "darkorchid3".

Let the $x$-axis range from 0 to 10, and let the $y$-axis range from 0 to 20.

Remember to include a main title and to properly label the axes.

**Solution**

```{r}
# Type your solution in here
```


## Exercise 2.6: Adjusting the shape of a point

Create a graph of a single circular point at $x = 7$ and $y = 12$, and use a character expansion factor of 3.5, an outline color of "darkred", an outline width of 2, and a fill color of "salmon1".

Use a `pch` value of 21 to obtain a circular point with a color fill.

Let the $x$-axis range from 0 to 10, and let the $y$-axis range from 0 to 20.

Remember to include a main title and to properly label the axes. 

**Solution**

```{r}
# Type your answer in here
```





## Exercise 2.7: Plotting multiple points

Using one call to the `plot()` function, plot these points:

| $X$ | $Y$ |
|-----|-----|
|  3  | 25  |
|  6  | 22  |
|  8  | 21  |
| 10  | 17  |
| 13  | 15  |

**Solution**


```{r}
# Type your answer in here
```




## Exercise 2.8: Modern Art


Express the poetic yearnings of your soul by creating modern art:

* Create a completely blank plot, with $x$ ranging from 0 to 10, and $y$ ranging from 0 to 10 as well. Make the plot square.

* In the upper left quadrant, draw an upright triangle colored red. Make it big!

* In the upper right quadrant, draw an inverted triangle with a black border, and a yellow fill. Make it the same size as the first triangle.

* In the lower left quadrant, draw a circle with a black border and a green fill. Make it the same size as the other objects.

* Finally, in the lower right quadrant draw a solid square colored blue.

**Solution**

```{r}
# Type your answer in here
```





# Solutions to the Exercises

## Exercise 2.1: Graphing a single point

First, create an empty plotting region with no data, where the $x$-values range from 0 to 10, and the $y$-values range from 0 to 8$.

Then use the `points()` function to draw a single point located at $x = 7$ and $y = 3$.

**Solution**

```{r}
plot(
    x = NULL,
    xlim = c(0, 10),
    ylim = c(0, 8),
    main = "Plot of a single point",
    xlab = "x",
    ylab = "y",
    las = 1
)

points(
    x = 7,
    y = 3.5
)
```


## Exercise 2.2: Annotating with text

Draw a graph with a single point located at $x = 7$ and $y = 3$.

Set the $x$-values for the plotting region to range from 0 to 10, and the $y$-values to range from 0 to 8$.

Then annotate this point with its coordinates using the `text()` function.

**Solution**

```{r}
plot(
    x = 7,
    y = 3,
    xlim = c(0, 10),
    ylim = c(0, 8),
    main = "Plot of a single point",
    xlab = "x",
    ylab = "y",
    las = 1
)

text(
    x = 7,
    y = 3.5,
    labels = "(7, 3)"
)
```


## Exercise 2.3: Adjusting the size of a point

Create a graph of the single point at $x = 7$ and $y = 12$, and use a character expansion factor of 3.5.

Let the $x$-axis range from 0 to 10, and let the $y$-axis range from 0 to 20.

Remember to include a main title and to properly label the axes.

**Solution**

Here's one approach:
```{r}
plot(
    x = 7,
    y = 12,
    xlim = c(0, 10),
    ylim = c(0, 20),
    main = "Graph of scaled single point",
    xlab = "x",
    ylab = "y",
    cex = 3.5,
    las = 1
)
```


Here's another approach:
```{r}
# First, create an empty plot with no data:

plot(
    x = NULL,
    xlim = c(0, 10),
    ylim = c(0, 20),
    main = "Graph of scaled single point",
    xlab = "x",
    ylab = "y",
    las = 1
)

# Then plot the point:

points(
    x = 7,
    y = 12,
    cex = 3.5
)

```




## Exercise 2.4: Adjusting the color of a point

Create a graph of the single point at $x = 7$ and $y = 12$, using a character expansion factor of 3.5 and a color of "hotpink3".

Let the $x$-axis range from 0 to 10, and let the $y$-axis range from 0 to 20.

Remember to include a main title and to properly label the axes.

**Solution**

Here's my solution:
```{r}
plot(
    x = 7,
    y = 12,
    xlim = c(0, 10),
    ylim = c(0, 20),
    main = "Graph of hotpink single point",
    xlab = "x",
    ylab = "y",
    cex = 3.5,
    col = "hotpink3",
    las = 1
)
```



## Exercise 2.5: Adjusting the shape of a point

Create a graph of the single point with a solid square shape at $x = 7$ and $y = 12$, using a character expansion factor of 3.5 and the color "darkorchid3".

Let the $x$-axis range from 0 to 10, and let the $y$-axis range from 0 to 20.

Remember to include a main title and to properly label the axes.


**Solution**

Here's my solution:
```{r}
plot(
    x = 7,
    y = 12,
    xlim = c(0, 10),
    ylim = c(0, 20),
    main = "Graph of scaled single point",
    xlab = "x",
    ylab = "y",
    cex = 3.5,
    col = "darkorchid3",
    pch = 15,
    las = 1
)
```



## Exercise 2.6: Adjusting the shape of a point

Create a graph of a single circular point at $x = 7$ and $y = 12$, and use a character expansion factor of 3.5, an outline color of "darkred", an outline width of 2, and a fill color of "salmon1".

Let the $x$-axis range from 0 to 10, and let the $y$-axis range from 0 to 20.

Remember to include a main title and to properly label the axes. 

**Solution**

```{r}
plot(
    x = 7,
    y = 12,
    xlim = c(0, 10),
    ylim = c(0, 20),
    main = "Graph of scaled single point",
    xlab = "x",
    ylab = "y",
    cex = 3.5,
    lwd = 2,
    col = "darkred",
    bg = "salmon1",
    pch = 21,
    las = 1
)
```







## Exercise 2.7: Plotting multiple points

Using one call to the `plot()` function, plot these points:

| $X$ | $Y$ |
|-----|-----|
|  3  | 25  |
|  6  | 22  |
|  8  | 21  |
| 10  | 17  |
| 13  | 15  |

**Solution**


First, let's create the $X$ and $Y$ vectors:
```{r}
x.vector <- c(3, 6, 8, 10, 13)

y.vector <- c(25, 22, 21, 17, 15)
```

Now we can plot all these values in just one call to the `plot()` function:
```{r}
plot(
    x.vector,
    y.vector,
    main = "Plotting multiple points at once",
    xlim = c(0, 15),
    ylim = c(0, 30),
    xlab = "x",
    ylab = "y",
    pch = 19,
    cex = 2,
    col = "darkslategray3",
    las = 1
)
```



## Exercise 2.8: Modern Art


Express the poetic yearnings of your soul by creating modern art:

* Create a completely blank plot, with $x$ ranging from 0 to 10, and $y$ ranging from 0 to 10 as well. Make the plot square.

* In the upper left quadrant, draw an upright triangle colored red. Make it big!

* In the upper right quadrant, draw an inverted triangle with a black border, and a yellow fill. Make it the same size as the same triangle.

* In the lower left quadrant, draw a circle with a black border and a green fill. Make it the same size as the other objects.

* Finally, in the lower right quadrant draw a solid square colored blue.

**Solution**

```{r}
# First, to make the image a square,
# we set the graphical parameter `pty`
# by assigning the value "s" to it:

par( pty = "s" )


# Next, create a completely blank plot

plot( x = NULL,
      xlim = c(0, 10),
      ylim = c(0, 10),
      axes = FALSE,
      xlab = "",
      ylab = ""
)


# Finally, draw the shapes:

points( 3, 7, pch = 17, cex = 10, col = "red" )
points( 7, 8, pch = 25, cex = 10, bg = "yellow", lwd = 2 )
points( 3, 3, pch = 21, cex = 10, bg = "green", lwd = 2 )
points( 7, 3, pch = 15, cex = 10, col = "blue" )

```






