---
title: "R Notebook"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


# Exercises



## Exercise 1.1: Logical operations

Perform a vectorized operation on the two logical vectors `first.logical.vector` and `second.logical.vector` by using the logical `or` operator denoted by `|`.

Can you predict what the resulting vector will be before you run the code?

**Solution**




## Exercise 1.2: Constructing a logical vector

Let's create a numeric vector:
```{r}
second.numeric.vector <-
    c( 6, 3, 1, 9, -2, 0, -7, 5, 8, -1)
```

Construct a logical vector which has the value `TRUE` in the locations where the corresponding element of `second.numeric.vector` is negative.

**Solution**





## Exercise 1.3: Logical functions

Let's start by creating two logical test vectors:
```{r}
## Example 24: Constructing two logical test vectors

first.logical.vector <-
    c(TRUE, TRUE, FALSE, TRUE, FALSE)

second.logical.vector <-
    c(TRUE, FALSE, TRUE, TRUE, FALSE)
```



Here's `second.logical.vector`:
```{r}
second.logical.vector
```

What is the result of calling the `all()` function with this vector?

What is the result of calling the `any()` function with this vector?

What is the result of calling the `which()` function with this vector?

Try to figure these out yourself before trying it in R.

**Solution**






# Solutions to the Exercises

## Exercise 1.1: Logical operations

Perform a vectorized operation on the two logical vectors `first.logical.vector` and `second.logical.vector` by using the logical `or` operator denoted by `|`.

Can you predict what the resulting vector will be before you run the code?

**Solution**

```{r}
first.logical.vector | second.logical.vector
```



## Exercise 1.2: Constructing a logical vector

Let's create a numeric vector:
```{r}
second.numeric.vector <-
    c( 6, 3, 1, 9, -2, 0, -7, 5, 8, -1)
```

Construct a logical vector which has the value `TRUE` in the locations where the corresponding element of `second.numeric.vector` is negative.

**Solution**

```{r}
second.numeric.vector < 0
```



## Exercise 1.3: Logical functions

Here's `second.logical.vector`:
```{r}
second.logical.vector
```

What is the result of calling the `all()` function with this vector?

What is the result of calling the `any()` function with this vector?

What is the result of calling the `which()` function with this vector?

Try to figure these out yourself before trying it in R:

**Solution**

Since at least one value in `second.logical.vector` is `FALSE`, the `all()` function returns `FALSE`:
```{r}
all( second.logical.vector )
```


Since at least one value in `second.logical.vector` is `TRUE`, the `all()` function returns `TRUE`:

```{r}
any( second.logical.vector )
```

Since `second.logical.vector` has the value `TRUE` in locations 1, 3, and 4, the `which()` function returns the values 1, 3, and 4:
```{r}
which( second.logical.vector)
```



