---
title: 'Lecture 2 Module 2: Variables'
subtitle: Exercises
output:
  pdf_document: default
  html_notebook: default
---


# Exercises



## Exercise 2.1: Initializing and modifying a variable

Initialize the variable `y` with the value 7, paying attention to the display in the Environment tab.

Then destructively modify it so that it is associated with the value -2.

**Solution**





## Exercise 2.2: Reporting numeric values

Report the value of `pi` using a `cat()` statement, displaying this value with 2 decimal places.

**Solution**





# Solution to the Exercises

## Exercise 2.1: Initializing and modifying a variable

Initialize the variable `y` with the value 7, paying attention to the display in the Environment tab.

Then destructively modify it so that it is associated with the value -2.

**Solution**

First, let's initialize the variable `y` with the value 7.
```{r}
y <- 7
```

Now let's destructively modify the variable, replacing the associated value with 2.
```{r}
y <- 2
```



## Exercise 2.2

Report the value of `pi` using a `cat()` statement, displaying this value with 2 decimal places.

**Solution**

```{r}
cat(
    "pi =",
    formatC(
        pi,
        format = "f",
        digits = 2
    )
)
```






