---
title: 'CSCI 5a Lecture 4 Module 3: Histograms'
subtitle: Exercises
output:
  html_document:
    df_print: paged
  pdf_document: default
---


```{r}
load( "Module 3 R Objects.Rdata")
```




# Exercises

## Exercise 3.1: Basic Histogram

Create a basic histogram for the values in `exercise.4.1.data`.

**Solution**




## Exercise 3.2: Basic Histogram

Construct a histogram for the numeric vector `exercise.4.1.data`. Include a main title and titles for the $x$-axis and $y$-axis, select a color for the bars, and explicitly specify the number of breaks.

**Solution**




# Solutions to the Exercises

## Exercise 3.1: Basic Histogram

Create a basic histogram for the values in `exercise.3.1.data`.

**Solution**

```{r}
hist( exercise.3.1.data )
```






## Exercise 3.2: Basic Histogram

Construct a histogram for the numeric vector `exercise.3.1.data`. Include a main title and titles for the $x$-axis and $y$-axis, select a color for the bars, and explicitly specify the number of breaks.

**Solution**

```{r}
hist(
    exercise.3.1.data,
    main = "Histogram of exercise.4.1.data",
    xlab = "x",
    ylab = "Frequency",
    col = "plum2",
    breaks = 50
)
```


