---
title: 'Week 1 Module 5: Curves'
subtitle: Exercises
output:
  pdf_document: default
  html_notebook: default
---

Let's clear the global computing environment:
```{r}
rm( list = ls() )
```



# Exercises for Week 1 Module 5: Curves


## Exercise 5.1: Cubic polynomial

Plot the function $f(x) = -2x^3 + 8x - 4$ for $x$ ranging from -3 to +3.

**Solution**

```{r}
# Type your answer here
```



## Exercise 5.2: Cubic polynomial with reference lines

Plot the function $f(x) = -2x^3 + 8x - 4$ for $x$ ranging from -3 to +3. Include the horizontal and vertical reference lines $y = 0$ and $x = 0$.

**Solution**

```{r}
# Type your solution here
```




## Exercise 5.3: Negative exponentials

Draw a graph of two functions:

* The function $f(x) = e^{-x}$.

* The function $g(x) = e^{-2x}$.

Draw horizontal and vertical reference lines, and annotate the curves with text.

**Solution**

```{r}
# Type your solution here
```





## Exercise 5.4: Negative exponentials with a legend

Draw a graph of two functions:

* The function $f(x) = e^{-x}$.

* The function $g(x) = e^{-2x}$.

Draw horizontal and vertical reference lines, annotate the curves with text, and include a legend.

**Solution**

```{r}
# Type your solution here
```





# Solutions to the Exercises

## Exercise 5.1: Cubic polynomial

Plot the function $f(x) = -2x^3 + 8x - 4$ for $x$ ranging from -3 to +3.

**Solution**

```{r}
curve(
    -2 * x^3 + 8 * x - 4,
    xlim = c(-3, 3),
    ylim = c(-30, 30),
    main = "Graph of f(x)",
    xlab = "x",
    ylab = "f(x)",
    lty = "solid",
    lwd = 3,
    col = "cyan4"
)
```



## Exercise 5.2: Cubic polynomial with reference lines

Plot the function $f(x) = -2x^3 + 8x - 4$ for $x$ ranging from -3 to +3. Include the horizontal and vertical reference lines $y = 0$ and $x = 0$.

**Solution**

```{r}
curve(
    -2 * x^3 + 8 * x - 4,
    xlim = c(-3, 3),
    ylim = c(-30, 30),
    main = "Graph of f(x)",
    xlab = "x",
    ylab = "f(x)",
    lty = "solid",
    lwd = 3,
    col = "cyan4"
)

# Draw the horizontal reference line:

segments(
    x0 = -3,
    y0 = 0,
    x1 = 3,
    y1 = 0,
    lty = "solid",
    lwd = 2,
    col = "gray50"
)


# Draw the vertical reference line:

segments(
    x0 = 0,
    y0 = -30,
    x1 = 0,
    y1 = 30,
    lty = "solid",
    lwd = 2,
    col = "gray50"
)

```




## Exercise 5.3: Negative exponentials

Draw a graph of two functions:

* The function $f(x) = e^{-x}$.

* The function $g(x) = e^{-2x}$.

Draw horizontal and vertical reference lines, and annotate the curves with text.

**Solution**

```{r}
plot(
    x = NULL,
    xlim = c(0, 3),
    ylim = c(0, 1),
    main = "Graph of negative exponential functions",
    xlab = "x",
    ylab = "f(x)"
)

# Now we can draw in the horizontal reference axis y = 0

segments(
    x0 = 0,
    y0 = 0,
    x1 = 3,
    y1 = 0,
    lwd = 2,
    lty = "solid",
    col = "gray50"
)

# Finally we can draw in the vertical reference axis x = 0

segments(
    x0 = 0,
    y0 = 0,
    x1 = 0,
    y1 = 1,
    lwd = 1,
    lty = "solid",
    col = "gray50"
)


curve(
    exp(-x),
    lwd = 3,
    lty = "solid",
    col = "blue",
    add = TRUE
)

text(
    x = 1,
    y = 0.45,
    labels = "f(x)"
)


curve(
    exp(-2 * x),
    lwd = 3,
    lty = "solid",
    col = "darkorchid3",
    add = TRUE
)

text(
    x = 0.5,
    y = 0.25,
    labels = "g(x)"
)

```


## Exercise 5.4: Negative exponentials with a legend

Draw a graph of two functions:

* The function $f(x) = e^{-x}$.

* The function $g(x) = e^{-2x}$.

Draw horizontal and vertical reference lines, annotate the curves with text, and include a legend.

**Solution**

```{r}
plot(
    x = NULL,
    xlim = c(0, 3),
    ylim = c(0, 1),
    main = "Graph of negative exponential functions",
    xlab = "x",
    ylab = "f(x)"
)

# Now we can draw in the horizontal reference axis y = 0

segments(
    x0 = 0,
    y0 = 0,
    x1 = 3,
    y1 = 0,
    lwd = 2,
    lty = "solid",
    col = "gray50"
)

# Finally we can draw in the vertical reference axis x = 0

segments(
    x0 = 0,
    y0 = 0,
    x1 = 0,
    y1 = 1,
    lwd = 1,
    lty = "solid",
    col = "gray50"
)


curve(
    exp(-x),
    lwd = 3,
    lty = "solid",
    col = "blue",
    add = TRUE
)

text(
    x = 1,
    y = 0.45,
    labels = "f(x)"
)


curve(
    exp(-2 * x),
    lwd = 3,
    lty = "solid",
    col = "darkorchid3",
    add = TRUE
)

text(
    x = 0.5,
    y = 0.25,
    labels = "g(x)"
)


legend(
    x = 2.4,
    y = 1,
    legend =
        c( "r = -1", "r = -2"),
    lty = "solid",
    lwd = 3,
    col =
        c( "blue", "darkorchid3")
)

```










