---
title: 'Week 6 Module 3: Tables'
subtitle: Exercises
output:
  pdf_document: default
  html_document:
    df_print: paged
---


Let's clear the environment:
```{r}
rm( list = ls() )
```

Let's load the objects for this module:
```{r}
load( "Module 3 R Objects.Rdata")

ls()
```




# Week 6 Module 3 Exercises


## Exercise 1: Constructing a frequency count table

The vector `one.week.cereal.brand.character.string.vector` contains data on cereal brands sold during one week.

First, construct a frequency count table for the cereal brands using this character string vector, save it in a variable, and display it directly.

Next, create a factor from this character string variable, and save it in a variable.

Then use this factor to construct a new table of the sample frequency count frequencies for this data.

**Solution**




## Exercise 2: Constructing a sample relative proportions table

Construct a sample relative proportions table for the cereal brands using the frequency count table you created in Exercise 1. Save this table in a variable, and display it directly.

**Solution**




## Exercise 3: Rounding the relative proportions table numbers

Use the `round()` function so that the numbers displayed in the relative proportions table have 3 decimal places.

**Solution**




## Exercise 4: Re-ordering the levels

Using character string indexing on the levels of the frequency count table that you created in Exercise 1 to create a new frequency count table displaying the brands in the order "Krispee Yummm!!", "Healthy Kale and Tofu", and "Sugar Bomz".

Then create a new factor with the levels pre-specified in the order "Krispee Yummm!!", "Healthy Kale and Tofu", and "Sugar Bomz". Save this factor in a variable, and use it to create another frequency count table, again displaying the brand names in this order.

Finally, sort the table in *ascending order*, so that the brand with the least sales is reported first, and the brand with the most sales is reported last.

**Solution**



## Exercise 5: Table and Pie Charts

Construct a pie chart of the relative proportions of the cereal brands using the frequency count table you constructed in Problem 1.

Then construct a pie chart of the relative proportions of the cereal brands using the frequency count table you constructed in Problem 1, displayed in descending order.

**Solution**





# Solutions to the Exercises


## Exercise 1: Constructing a frequency count table

The vector `one.week.cereal.brand.character.string.vector` contains data on cereal brands sold during one week.

First, construct a frequency count table for the cereal brands using this character string vector, save it in a variable, and display it directly.

Next, create a factor from this character string variable, and save it in a variable.

Then use this factor to construct a new table of the sample frequency count frequencies for this data.

**Solution**

Here's the table constructed from the character string vector:
```{r}
one.week.cereal.brand.frequency.count.table <-
  table(
    one.week.cereal.brand.character.string.vector
    )

one.week.cereal.brand.frequency.count.table
```

Next we can create the factor:
```{r}
one.week.cereal.brand.factor <-
  factor(
    one.week.cereal.brand.character.string.vector
  )
```

And finally we can construct another frequency count table:
```{r}
one.week.cereal.brand.frequency.count.table <-
  table(
    one.week.cereal.brand.factor
  )

one.week.cereal.brand.frequency.count.table
```



## Exercise 2: Constructing a sample relative proportions table

Construct a sample relative proportions table for the cereal brands using the frequency count table you created in Exercise 1. Save this table in a variable, and display it directly.

**Solution**

```{r}
one.week.cereal.brand.relative.proportions.table <-
  proportions(
    one.week.cereal.brand.frequency.count.table
  )

one.week.cereal.brand.relative.proportions.table
```



## Exercise 3: Rounding the relative proportions table numbers

Use the `round()` function so that the numbers displayed in the relative proportions table have 3 decimal places.

**Solution**

```{r}
round(
  one.week.cereal.brand.relative.proportions.table,
  digits = 3
)
```



## Exercise 4: Re-ordering the levels

Using character string indexing on the levels of the frequency count table that you created in Exercise 1 to create a new frequency count table displaying the brands in the order "Krispee Yummm!!", "Healthy Kale and Tofu", and "Sugar Bomz".

Then create a new factor with the levels pre-specified in the order "Krispee Yummm!!", "Healthy Kale and Tofu", and "Sugar Bomz". Save this factor in a variable, and use it to create another frequency count table, again displaying the brand names in this order.

Finally, sort the table in *ascending order*, so that the brand with the least sales is reported first, and the brand with the most sales is reported last.

**Solution**

```{r}
reordered.one.week.cereal.brand.frequency.count.table <-
  one.week.cereal.brand.frequency.count.table[
    c(
      "Krispee Yummm!!",
      "Healthy Kale and Tofu",
      "Sugar Bomz"
    )
  ]
```

```{r}
reordered.one.week.cereal.brand.factor <-
  factor(
    one.week.cereal.brand.character.string.vector,
    levels =
      c(
        "Krispee Yummm!!",
        "Healthy Kale and Tofu",
        "Sugar Bomz"
      )
  )

table(
  reordered.one.week.cereal.brand.factor
)
```

```{r}
ascending.order.one.week.cereal.brand.frequency.count.table <-
  sort( one.week.cereal.brand.frequency.count.table )

ascending.order.one.week.cereal.brand.frequency.count.table
```


## Exercise 5

Construct a pie chart of the relative proportions of the cereal brands using the frequency count table you constructed in Problem 1.

Then construct a pie chart of the relative proportions of the cereal brands using the frequency count table you constructed in Problem 1, displayed in decreasing order.

**Solution**

```{r}
pie(
  sort(
    one.week.cereal.brand.frequency.count.table,
    decreasing = TRUE
  ),
  main = "Pie chart of frequency counts for cereal brand sales",
  clockwise = TRUE
)
```



